---
title: "BayesianOrdinationRegressionAnaLysis"
author: "P Harnay"
date: "2025-11-24"
output: html_document
---

Joint species distribution model approach
Install and load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# download and install jags at https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/Mac%20OS%20X/

library(tidyverse)
library(rjags)
library(R2jags)
library(boral)
#Hui, F. K. C. (2016). boral—Bayesian ordination and regression analysis of multivariate abundance data in R. Methods in Ecology and Evolution, 7, 744–750.
library(dplyr)
library(reshape2)
library(geosphere)
library(dplyr)
library(ggplot2)
library(scales)


```

load and visualize coral density data
```{r}

data <- read.csv("Data/density.csv")%>%
  select("Site","Species","LifeStage","mean_density", "sd_density")

str(data)

data$LifeStage <- factor(
  data$LifeStage,
  levels = c("0-2cm", ">2cm-5cm", ">5cm-14cm", ">14cm")
)

data$Site <- factor(
  data$Site,
  levels = c("S2", "S3", "S4", "S5", "S6", "S9", "S11", "S12", "S14", "S15", "S18" )
)

unscaled.density <- ggplot(data, aes(x = Site,
                 y = mean_density,
                 fill = Species)) +   # <-- map Species to fill
  geom_col() +
  geom_errorbar(aes(ymin = mean_density - sd_density,
                    ymax = mean_density + sd_density),
                width = 0.2) +
  facet_grid(Species ~ LifeStage) +
  scale_fill_manual(values = c(
    "P. tuahiniensis" = "#D55E00",
    "P. meandrina"    = "#0072B2",
    "P. verrucosa"    = "#E69F00",
    "P. grandis"      = "#56B4E9",
    "P. effusa"       = "#009E73",
    "P. acuta"        = "#e63946"
  )) +
  labs(x = "Site",
       y = expression("Density (colonies m"^{-2}*")"),
       fill = "Species") +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    strip.background = element_rect(fill = "grey90")
  )
unscaled.density

scaled.density <- ggplot(data, aes(x = Site,
                 y = mean_density,
                 fill = Species)) +   # <-- map Species to fill
  geom_col() +
  geom_errorbar(aes(ymin = mean_density - sd_density,
                    ymax = mean_density + sd_density),
                width = 0.2) +
  facet_grid(Species ~ LifeStage, scales="free") +
  scale_fill_manual(values = c(
    "P. tuahiniensis" = "#D55E00",
    "P. meandrina"    = "#0072B2",
    "P. verrucosa"    = "#E69F00",
    "P. grandis"      = "#56B4E9",
    "P. effusa"       = "#009E73",
    "P. acuta"        = "#e63946"
  )) +
  labs(x = "Site",
       y = expression("Density (colonies m"^{-2}*")"),
       fill = "Species") +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    strip.background = element_rect(fill = "grey90")
  )

scaled.density

ggsave(
  filename = "Output/scaled_density_plot.png",
  plot = scaled.density,
  device = "png",
  width = 8,     # adjust width in inches
  height = 9,     # adjust height in inches
  units = "in"
)

```

1. Response matrix Y: coral density by species × size class × site
Reshape data to make a site × (species-sizeclass) matrix
```{r}
coral_long <- data

coral_wide <- coral_long %>%
  mutate(sp_sz = paste(Species, LifeStage, sep = "_")) %>%
  select(Site, sp_sz, mean_density) %>%
  pivot_wider(names_from = sp_sz,
              values_from = mean_density,
              values_fill = 0)

# response matrix Y
Y <- coral_wide %>%
  column_to_rownames("Site") %>%
  as.matrix()


```

2. Site table sites: PH locations + shore
list site locations and assign shore
```{r}
sites <- tribble(
  ~long,        ~lat,         ~site.name,
  -149.8764837, -17.48523962, "S1",
  -149.8526452, -17.48223933, "S2",
  -149.8344367, -17.47810177, "S3",
  -149.8076461, -17.47569278, "S4",
  -149.7835691, -17.47304608, "S5",
  -149.7614724, -17.50575957, "S6",
  -149.7845548, -17.56535312, "S9",
  -149.804473,  -17.60276752, "S11",
  -149.8478059, -17.59508584, "S12",
  -149.8837824, -17.56500936, "S14",
  -149.9007712, -17.54654008, "S15",
  -149.9239404, -17.49378121, "S18"
)

sites <- sites %>%
  mutate(
    shore = case_when(
      site.name %in% c("S1","S2","S3","S4","S5") ~ "North",
      site.name %in% c("S6","S9","S11") ~ "East",
      site.name %in% c("S12","S14","S15", "S18") ~ "West"
    ),
    shore = factor(shore, levels = c("West","North","East"))
  )

sites <- sites %>%
  filter(site.name %in% rownames(Y)) %>%
  arrange(match(site.name, rownames(Y)))

stopifnot(all(sites$site.name == rownames(Y)))  # sanity check

```

3. Wave forcing: fore-reef energy, 3-shore resolution
Wave Forcing
3.1 wave download function
```{r}
options(HTTPUserAgent = "EDI_CodeGen")

read_mcr_wave <- function(url) {
  tmp <- tempfile()
  # Try curl first (needed for EDI sometimes)
  try(
    download.file(
      url, tmp,
      method = "curl",
      extra  = paste0(' -A "', getOption("HTTPUserAgent"), '"')
    ),
    silent = TRUE
  )

  # Fallback to default method if curl failed
  if (is.na(file.size(tmp))) {
    download.file(url, tmp, method = "auto")
  }

  # Read the data: skip the first line (metadata), use the second as header
  dat <- read.csv(
    tmp,
    header = TRUE,
    skip   = 0,
    sep    = ",",
    quote  = '"',
    check.names = TRUE,
    stringsAsFactors = FALSE
  )

  unlink(tmp)

  # If the file uses time_local / time_utc columns, parse them
  if ("time_local" %in% names(dat)) {
    suppressWarnings({
      dat$time_local <- as.POSIXct(dat$time_local, tz = "Pacific/Midway") # adjust TZ if needed
    })
  }
  if ("time_utc" %in% names(dat)) {
    suppressWarnings({
      dat$time_utc <- as.POSIXct(dat$time_utc, tz = "UTC")
    })
  }

  dat
}


```

3.2 Download data
```{r}
# Package ID: knb-lter-mcr.30.44 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Ocean Currents and Biogeochemistry: salinity, temperature and current at CTD and ADCP mooring FOR01 from 2004 ongoing.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Libe Washburn - Moorea Coral Reef LTER 
# Metadata Provider:    - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu
wave_lter1 <- read_mcr_wave(
  "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/30/44/e29d12699f7556d886c3b3c275370b42")

# Package ID: knb-lter-mcr.31.45 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Ocean Currents and Biogeochemistry: salinity, temperature and current at CTD and ADCP mooring FOR04 from 2004 ongoing.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Libe Washburn - Moorea Coral Reef LTER 
# Metadata Provider:    - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu
wave_lter4 <- read_mcr_wave("https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/31/45/e29d12699f7556d886c3b3c275370b42")

# Package ID: knb-lter-mcr.32.41 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Ocean Currents and Biogeochemistry: salinity, temperature and current at CTD and ADCP mooring FOR05 from 2005 ongoing.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Libe Washburn - Moorea Coral Reef LTER 
# Metadata Provider:    - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu
wave_lter5 <- read_mcr_wave("https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/32/41/e29d12699f7556d886c3b3c275370b42")


wave_lter1 <- wave_lter1 %>%
  mutate(
    # Replace -999 with NA for the key wave variables
    wave_ht_adcp_m = na_if(wave_ht_adcp_m, -999),
    wave_period_adcp_sec = na_if(wave_period_adcp_sec, -999),
    wave_power_adcp = na_if(wave_power_adcp, -999)
  )

str(wave_lter1)
names(wave_lter1)
head(wave_lter1)

wave_lter4 <- wave_lter4 %>%
  mutate(
    # Replace -999 with NA for the key wave variables
    wave_ht_adcp_m = na_if(wave_ht_adcp_m, -999),
    wave_period_adcp_sec = na_if(wave_period_adcp_sec, -999),
    wave_power_adcp = na_if(wave_power_adcp, -999)
  )

str(wave_lter4)
names(wave_lter4)
head(wave_lter4)

wave_lter5 <- wave_lter5 %>%
  mutate(
    # Replace -999 with NA for the key wave variables
    wave_ht_adcp_m = na_if(wave_ht_adcp_m, -999),
    wave_period_adcp_sec = na_if(wave_period_adcp_sec, -999),
    wave_power_adcp = na_if(wave_power_adcp, -999)
  )

str(wave_lter5)
names(wave_lter5)
head(wave_lter5)

#filter to dates after > 2017
wave_lter1 <- wave_lter1 %>%
  filter(time_local > as.Date("2016-12-31"))
wave_lter4 <- wave_lter4 %>%
  filter(time_local > as.Date("2016-12-31"))
wave_lter5 <- wave_lter5 %>%
  filter(time_local > as.Date("2016-12-31"))

```

3.3 Function for Calculating Wave Power
```{r}
#Calculate P = wave energy flux (kW m–1)
#From Dean & Dalrymple (1991)
#P = wave energy flux (W m⁻¹)
#ρ=1028 kg m⁻³ (seawater density)
#g=9.8 m s⁻²
#Hsig = significant wave height (m)
#T = dominant wave period (s)

rho <- 1028     # seawater density
g <- 9.8        # gravity


process_wave_site <- function(dat, shore_label, start_date = "2017-01-01") {

  dat_clean <- dat %>%
    mutate(
      wave_ht_adcp_m       = na_if(wave_ht_adcp_m, -999),
      wave_period_adcp_sec = na_if(wave_period_adcp_sec, -999),
      wave_power_adcp      = na_if(wave_power_adcp, -999),
      Hsig = wave_ht_adcp_m,
      T    = wave_period_adcp_sec,
      P_wm  = (rho * g^2 * Hsig^2 * T) / (64 * pi),
      P_kwm = P_wm / 1000
    )

  daily <- dat_clean %>%
    mutate(date = as.Date(time_local)) %>%
    group_by(date) %>%
    summarise(
      wave_power_adcp   = mean(wave_power_adcp,   na.rm = TRUE),
      wave_ht_adcp_m    = mean(wave_ht_adcp_m,    na.rm = TRUE),
      wave_period_adcp  = mean(wave_period_adcp_sec, na.rm = TRUE),
      P_kwm             = mean(P_kwm, na.rm = TRUE),
      .groups = "drop"
    )

  clim <- daily %>%
    filter(date >= as.Date(start_date)) %>%
    summarise(
      P_mean = mean(P_kwm, na.rm = TRUE),
      P_p95  = quantile(P_kwm, 0.95, na.rm = TRUE),
      P_max  = max(P_kwm, na.rm = TRUE),
      P_sd   = sd(P_kwm, na.rm = TRUE)
    ) %>%
    mutate(shore = shore_label)

  list(daily = daily, clim = clim)
}

```

3.4 Calculate Wave Power
```{r}

#LTER_1
res_lter1 <- process_wave_site(wave_lter1, shore_label = "North")
wave_lter1_daily <- res_lter1$daily
wave_lter1_clim  <- res_lter1$clim

#LTER_4
res_lter4 <- process_wave_site(wave_lter4, shore_label = "East")
wave_lter4_daily <- res_lter4$daily
wave_lter4_clim  <- res_lter4$clim

#LTER_5
res_lter5 <- process_wave_site(wave_lter5, shore_label = "West")
wave_lter5_daily <- res_lter5$daily
wave_lter5_clim  <- res_lter5$clim



```

3.5 Plot daily P_kwm for all 3 sites on the same axes
```{r}
wave_daily_all <- bind_rows(
  wave_lter1_daily %>% mutate(site = "LTER_1 (North)"),
  wave_lter4_daily %>% mutate(site = "LTER_4 (East)"),
  wave_lter5_daily %>% mutate(site = "LTER_5 (West)")
)

waves <- ggplot(wave_daily_all, aes(x = date, y = P_kwm)) +
  geom_line(linewidth = 0.3) +
  facet_wrap(~ site, nrow = 1, scales = "fixed") +  # same y-axis across facets
  labs(
    #title = "Wave energy flux (kW m^-1) – Fore Reef Stations",
    x = "Date",
    y = expression("Wave energy flux (kW m"^{-1}*")")
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold")
  )

ggsave(
  filename = "Output/waves_plot.png",
  plot = waves,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)


```

3.6 Build the shore-level wave table for the JSDM
```{r}

wave_shore <- bind_rows(
  wave_lter1_clim,
  wave_lter4_clim,
  wave_lter5_clim
) %>%
  select(shore, P_mean, P_p95, P_max, P_sd)

wave_shore

sites_wave <- sites %>%
  left_join(wave_shore, by = "shore")

str(sites_wave)

```

4. Lagoon temperature: site-level from SBE56 array
Temperature

4.1 Downloading temperature data
```{r}
# Package ID: knb-lter-mcr.1045.6 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Island-Wide Lagoon Benthic Water Temperature, ongoing since 2021.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Tom Adam - Moorea Coral Reef LTER 
# Data set creator:  James Hench - Moorea Coral Reef LTER 
# Data set creator:  Andrew Brooks - Moorea Coral Reef LTER 
# Metadata Provider:    - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu
# Stylesheet v2.15 for metadata conversion into program: John H. Porter, Univ. Virginia, jporter@virginia.edu      
# Uncomment the following lines to have R clear previous work, or set a working directory
# rm(list=ls())      

# setwd("C:/users/my_name/my_dir")       

options(HTTPUserAgent = "EDI_CodeGen")

## ---------- 1. Daily lagoon temperature (rename dt1 → temp_daily) ----------

temp_url  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1045/6/7a8090507e95d40ba98d837388f7c506"
temp_file <- tempfile()

try(
  download.file(
    temp_url,
    temp_file,
    method = "curl",
    extra = paste0(' -A "', getOption("HTTPUserAgent"), '"')
  )
)

if (is.na(file.size(temp_file))) {
  download.file(temp_url, temp_file, method = "auto")
}

lagoon_temp_daily <- read.csv(
  temp_file,
  header   = FALSE,
  skip     = 1,
  sep      = ",",
  quote    = '"',
  col.names = c(
    "site",
    "date_local",
    "min_temp_c",
    "max_temp_c",
    "median_temp_c",
    "mean_temp_c"
  ),
  check.names = TRUE
)

unlink(temp_file)

## Type fixes / parsing (no attach)

lagoon_temp_daily <- lagoon_temp_daily %>%
  mutate(
    site = as.factor(site),
    # parse date
    date_local = as.Date(date_local, format = "%Y-%m-%d"),
    # numeric temps
    across(
      c(min_temp_c, max_temp_c, median_temp_c, mean_temp_c),
      ~ suppressWarnings(as.numeric(.))
    )
  )

str(lagoon_temp_daily)
summary(lagoon_temp_daily$site)
summary(lagoon_temp_daily$mean_temp_c)

#plot mean data to look for outliers
ggplot(lagoon_temp_daily, aes(x = date_local, y = mean_temp_c)) +
  geom_line(color = "steelblue", linewidth = 0.3) +
  facet_wrap(~ site, scales = "free_y", ncol = 7) +
  theme_bw(base_size = 9) +
  labs(
    title = "Daily Mean Temperature at All Lagoon Sites",
    x = "Date",
    y = "Mean Temp (°C)"
  )

#plot min data to look for outliers
ggplot(lagoon_temp_daily, aes(x = date_local, y = min_temp_c)) +
  geom_line(color = "steelblue", linewidth = 0.3) +
  facet_wrap(~ site, scales = "free_y", ncol = 7) +
  theme_bw(base_size = 9) +
  labs(
    title = "Daily Minimum Temperature at All Lagoon Sites",
    x = "Date",
    y = "Minimum Temp (°C)"
  )

#plot max data to look for outliers
ggplot(lagoon_temp_daily, aes(x = date_local, y = max_temp_c)) +
  geom_line(color = "steelblue", linewidth = 0.3) +
  facet_wrap(~ site, scales = "free_y", ncol = 7) +
  theme_bw(base_size = 9) +
  labs(
    title = "Daily Maximum Temperature at All Lagoon Sites",
    x = "Date",
    y = "Maximum Temp (°C)"
  )



## ---------- 2. Site–filename lookup (rename dt2 → temp_files) ----------

temp_meta_url  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1045/6/aa66754bc0fec207b813ee970389ad93"
temp_meta_file <- tempfile()

try(
  download.file(
    temp_meta_url,
    temp_meta_file,
    method = "curl",
    extra = paste0(' -A "', getOption("HTTPUserAgent"), '"')
  )
)

if (is.na(file.size(temp_meta_file))) {
  download.file(temp_meta_url, temp_meta_file, method = "auto")
}

lagoon_temp_files <- read.csv(
  temp_meta_file,
  header   = FALSE,
  skip     = 1,
  sep      = ",",
  quote    = '"',
  col.names = c("site_code", "filename"),
  check.names = TRUE
)

unlink(temp_meta_file)

lagoon_temp_files <- lagoon_temp_files %>%
  mutate(
    site_code = as.factor(site_code),
    filename  = as.factor(filename)
  )

str(lagoon_temp_files)
summary(lagoon_temp_files$site_code)


```

4.2 Summarize temperature data
```{r}
lagoon_temp_summary <- lagoon_temp_daily %>%
  group_by(site) %>%
  summarise(
    mean_temp        = mean(mean_temp_c, na.rm = TRUE),
    median_temp      = median(median_temp_c, na.rm = TRUE),
    max_temp         = max(max_temp_c, na.rm = TRUE),
    p95_temp         = quantile(max_temp_c, 0.95, na.rm = TRUE),
    sd_temp          = sd(mean_temp_c, na.rm = TRUE),
    mean_daily_range = mean(max_temp_c - min_temp_c, na.rm = TRUE),
    freq_above_30C   = mean(max_temp_c > 30, na.rm = TRUE),
    n_days           = sum(!is.na(mean_temp_c)),
    .groups = "drop"
  )

lagoon_temp_summary

lagoon_coords <- tribble(
  ~site,   ~long,      ~lat,
  # Backbone B01–B35
  "B01",   -149.89,    -17.485,
  "B02",   -149.879,   -17.484,
  "B03",   -149.868,   -17.487,
  "B04",   -149.85,    -17.482,
  "B05",   -149.833,   -17.479,
  "B06",   -149.817,   -17.475,
  "B07",   -149.806,   -17.475,
  "B08",   -149.793,   -17.475,
  "B09",   -149.784,   -17.473,
  "B10",   -149.761,   -17.503,
  "B10.5", -149.763,   -17.508,
  "B11",   -149.765,   -17.520,
  "B12",   -149.765,   -17.530,
  "B13",   -149.764,   -17.535,
  "B14",   -149.767,   -17.541,
  "B15",   -149.783,   -17.561,
  "B16",   -149.787,   -17.569,
  "B17",   -149.789,   -17.580,
  "B18",   -149.791,   -17.587,
  "B19",   -149.799,   -17.599,
  "B20",   -149.808,   -17.603,
  "B21",   -149.817,   -17.603,
  "B22",   -149.827,   -17.603,
  "B23",   -149.836,   -17.599,
  "B24",   -149.853,   -17.593,
  "B25",   -149.861,   -17.590,
  "B26",   -149.867,   -17.585,
  "B27",   -149.882,   -17.568,
  "B28",   -149.885,   -17.562,
  "B29",   -149.892,   -17.555,
  "B30",   -149.897,   -17.550,
  "B31",   -149.902,   -17.545,
  "B32",   -149.913,   -17.535,
  "B33",   -149.922,   -17.513,
  "B34",   -149.924,   -17.504,
  "B35",   -149.923,   -17.495,

  # North transect N01–N05
  "N01",   -149.84593, -17.48063,
  "N02",   -149.84624, -17.48011,
  "N03",   -149.84657, -17.47915,
  "N04",   -149.84544, -17.48146,
  "N05",   -149.84500, -17.48200,

  # West transect W01–W05
  "W01",   -149.86800, -17.58630,
  "W02",   -149.86721, -17.58585,
  "W04",   -149.86654, -17.58431,
  "W05",   -149.86600, -17.58300,

  # East transect E01–E04
  "E01",   -149.76600, -17.54090,
  "E02",   -149.76600, -17.54110,
  "E03",   -149.76800, -17.54030,
  "E04",   -149.76800, -17.53970
)

lagoon_temp_full <- lagoon_temp_summary %>%
  left_join(lagoon_coords, by = "site")

```

4.3 Map temperature data to coral sites
```{r}
ph_coords  <- as.matrix(sites[, c("long", "lat")])
temp_coords <- as.matrix(lagoon_temp_full[, c("long", "lat")])

dist_mat_temp <- distm(ph_coords, temp_coords, fun = distHaversine)

nn_temp_idx <- apply(dist_mat_temp, 1, which.min)
dist_to_temp_m <- apply(dist_mat_temp, 1, min)

#join
sites_temp <- sites %>%
  mutate(
    temp_site         = lagoon_temp_full$site[nn_temp_idx],
    temp_long         = lagoon_temp_full$long[nn_temp_idx],
    temp_lat          = lagoon_temp_full$lat[nn_temp_idx],
    mean_temp_br      = lagoon_temp_full$mean_temp[nn_temp_idx],
    median_temp_br    = lagoon_temp_full$median_temp[nn_temp_idx],
    max_temp_br       = lagoon_temp_full$max_temp[nn_temp_idx],
    p95_temp_br       = lagoon_temp_full$p95_temp[nn_temp_idx],
    daily_range_br    = lagoon_temp_full$mean_daily_range[nn_temp_idx],
    sd_temp_br        = lagoon_temp_full$sd_temp[nn_temp_idx],
    freq_above_30C_br = lagoon_temp_full$freq_above_30C[nn_temp_idx],
    dist_to_temp_m    = dist_to_temp_m
  )

#inspect
sites_temp %>%
  select(site.name, shore, temp_site, dist_to_temp_m,
         mean_temp_br, max_temp_br, p95_temp_br, daily_range_br)

str(sites_temp)

```
View temperature data daily
```{r}
## 1. Extract mapping: coral site.name → temp_site
temp_map <- sites_temp %>%
  select(site.name, temp_site)

## 2. Filter lagoon_temp_daily to only the 11 temp sites
lagoon_temp_coral <- lagoon_temp_daily %>%
  filter(site %in% unique(temp_map$temp_site)) %>%
  mutate(daily_range_c = max_temp_c - min_temp_c)

## 3. Join so each row knows its coral site.name
lagoon_temp_joined <- lagoon_temp_coral %>%
  left_join(temp_map, by = c("site" = "temp_site")) %>%
  # reorder levels if you want S11 → S36 in order
  mutate(site.name = factor(site.name, levels = sort(unique(site.name))))

## ---- CHECK RESULT
# head(lagoon_temp_joined)

temp_max <- ggplot(
  lagoon_temp_joined %>% filter(!is.na(max_temp_c)),
  aes(x = date_local, y = max_temp_c, color = site.name)
) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_color_viridis_d(option = "turbo") +
  theme_bw(base_size = 12) +
  labs(
    title = "Daily Maximum Temperature at Coral Sites",
    x = "Date",
    y = "Maximum Temperature (°C)",
    color = "Coral Site"
  )


temp_range <- ggplot(
  lagoon_temp_joined %>% filter(!is.na(daily_range_c)),
  aes(x = date_local, y = daily_range_c, color = site.name)
) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_color_viridis_d(option = "magma") +
  theme_bw(base_size = 12) +
  labs(
    title = "Daily Temperature Range at Coral Sites",
    x = "Date",
    y = "Temperature Range (°C)",
    color = "Coral Site"
  )

ggsave(
  filename = "Output/temp_max_plot.png",
  plot = temp_max,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)

ggsave(
  filename = "Output/temp_range_plot.png",
  plot = temp_range,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)

```

5. Benthic cover: backreef photoquadrats, block-level

5.1 Download benthic cover data and clean
```{r}
# # Package ID: knb-lter-mcr.1038.14 Cataloging System:https://pasta.edirepository.org.
# # Data set title: MCR LTER: Coral Reef: Long-term Community Dynamics: Backreef (Lagoon) Corals Annual Survey, ongoing since 2005.
# # Data set creator:    - Moorea Coral Reef LTER 
# # Data set creator:  Peter Edmunds - Moorea Coral Reef LTER 
# # Metadata Provider:  Hillary Krumbholz - MCR LTER 
# # Contact:    - Information Manager MCR LTER  - hkrumbholz@ucsb.edu
# 
# options(HTTPUserAgent = "EDI_CodeGen")
# 	      
# # URL for table 1 (simple cover: coral / macroalgae / sand / millepora / ctb)
# benthic_simple_url <- 
#   "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1038/14/b7ca625d17a5630441af8c229cc0ab92"
# 
# benthic_simple_file <- tempfile()
# 
# try(
#   download.file(
#     benthic_simple_url, 
#     benthic_simple_file,
#     method = "curl",
#     extra = paste0(' -A "', getOption("HTTPUserAgent"), '"')
#   ),
#   silent = TRUE
# )
# 
# if (is.na(file.size(benthic_simple_file))) {
#   download.file(benthic_simple_url, benthic_simple_file, method = "auto")
# }
# 
# # Read cleanly
# benthic_simple_raw <- read_csv(
#   benthic_simple_file,
#   skip = 1,
#   col_names = c(
#     "year", "site", "habitat", "depth", "quad", "patch",
#     "coral", "macroalgae", "sand", "millepora", "ctb",
#     "sub_squares", "note"
#   ),
#   na = c("na", "NA", "BW", "")
# )
# 
# unlink(benthic_simple_file)
# 
# benthic_simple <- benthic_simple_raw %>%
#   mutate(
#     site    = as.factor(site),
#     habitat = as.factor(habitat),
#     quad    = as.factor(quad),
#     patch   = as.factor(patch),
#     note    = as.factor(note),
#     depth   = suppressWarnings(as.numeric(depth)),
#     across(where(is.character))  # ensures numeric cols clean
#   )

```
5.1 Importing data from benthics cover (site characterization)

```{r}
benthic_data <- read.csv("Data/20260226_Quadrat_Benthos_RA.csv")

str(benthic_data)

#split the quadrats name in different part to show the quadrats number, site and transect 
benthics_data_split <- benthic_data %>%
   separate(
    Name,
    into = c("Quadrat", "Site", "Transect"),
    sep = "_"
  ) %>%
  mutate(
    Transect = str_remove(Transect, "\\.JPG")
  )

#select only transect 1 to 3 and exclude 4 for 6 
benthic_data_T123 <- benthics_data_split %>%
  filter(!(
    (Site == "S1" & Transect == "T4") |
    (Site == "S1" & Transect == "T5") |
    (Site == "S1" & Transect == "T6") |
    (Site == "S2" & Transect == "T4") |
    (Site == "S2" & Transect == "T5") |
    (Site == "S2" & Transect == "T6") |
    (Site == "S3" & Transect == "T4") |
    (Site == "S3" & Transect == "T5") |
    (Site == "S3" & Transect == "T6") |
    (Site == "S4" & Transect == "T4") |
    (Site == "S4" & Transect == "T5") |
    (Site == "S4" & Transect == "T6") |
    (Site == "S4" & Transect == "T6") |
    (Site == "S5" & Transect == "T4") |
    (Site == "S5" & Transect == "T5") |
    (Site == "S5" & Transect == "T6") |
    (Site == "S6" & Transect == "T4") |
    (Site == "S6" & Transect == "T5") |
    (Site == "S6" & Transect == "T6") |
    (Site == "S9" & Transect == "T4") |
    (Site == "S9" & Transect == "T5") |
    (Site == "S9" & Transect == "T6") |
    (Site == "S11" & Transect == "T4") |
    (Site == "S11" & Transect == "T5") |
    (Site == "S11" & Transect == "T6") |
    (Site == "S12" & Transect == "T4") |
    (Site == "S12" & Transect == "T5") |
    (Site == "S12" & Transect == "T6") |
    (Site == "S14" & Transect == "T4") |
    (Site == "S14" & Transect == "T5") |
    (Site == "S14" & Transect == "T6") |
    (Site == "S15" & Transect == "T4") |
    (Site == "S15" & Transect == "T5") |
    (Site == "S15" & Transect == "T6") |
    (Site == "S18" & Transect == "T4") |
    (Site == "S18" & Transect == "T5") |
    (Site == "S18" & Transect == "T6") |
    (is.na(Site) & Transect == "T2")
  ))


```
5.2 Process downloaded benthic cover data and map to sites
```{r}
# 1) Filter to backreef/lagoon habitat and summarise by site
benthic_br_site <- benthic_simple %>%
  # keep lagoon/backreef habitats; tweak this if you see slightly different labels
  filter(year > 2016,
          grepl("back", habitat, ignore.case = TRUE) | 
         grepl("lagoon", habitat, ignore.case = TRUE)) %>%
  mutate(
    # Make sure site is character, then create LTER-style ID
    lter_site = as.character(site)
  ) %>%
  group_by(lter_site) %>%
  summarise(
    coral_cover      = mean(coral,      na.rm = TRUE),
    macroalgae_cover = mean(macroalgae, na.rm = TRUE),
    sand_cover       = mean(sand,       na.rm = TRUE),
    millepora_cover  = mean(millepora,  na.rm = TRUE),
    ctb_cover        = mean(ctb,        na.rm = TRUE),
    .groups = "drop"
  )

benthic_br_site

benthic_long <- benthic_br_site %>%
  pivot_longer(
    cols = -lter_site,
    names_to = "cover_type",
    values_to = "percent_cover"
  )

benthic_cover <- ggplot(benthic_long, aes(x = lter_site, y = percent_cover, fill = cover_type)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    #title = "Benthic Cover Composition by Site",
    x = "LTER Site",
    y = "Percent Cover",
    fill = "Benthic Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
benthic_cover

ggsave(
  filename = "Output/benthic_cover_plot.png",
  plot = benthic_cover,
  device = "png",
  width = 5,     # adjust width in inches
  height = 3,     # adjust height in inches
  units = "in"
)



backreef_coords <- tribble(
  ~site,    ~long,        ~lat,
  "LTER_1", -149.8425765, -17.47823787,
  "LTER_2", -149.8033863, -17.47700229,
  "LTER_3", -149.7643003, -17.51396455,
  "LTER_4", -149.7696695, -17.54318109,
  "LTER_5", -149.8714356, -17.57932994,
  "LTER_6", -149.9210026, -17.52112157
)

# Join coordinates with benthic means
backreef_benthic <- backreef_coords %>%
  left_join(benthic_br_site, by = c("site" = "lter_site"))

## 1. Coordinate matrices: PH sites and backreef sites ----

# Make sure 'sites' has columns: site.name, long, lat
# And 'backreef_benthic' has: site, long, lat, coral_cover, macroalgae_cover, ctb_cover, sand_cover, millepora_cover

ph_coords       <- as.matrix(sites[, c("long", "lat")])
backreef_coords <- as.matrix(backreef_benthic[, c("long", "lat")])

## 2. Distance matrix + nearest neighbor index ----

# Distance matrix (meters): rows = PH sites, cols = LTER backreef sites
dist_mat <- distm(ph_coords, backreef_coords, fun = distHaversine)

# Index of nearest backreef site for each PH site
nn_idx <- apply(dist_mat, 1, which.min)

# Minimum distance (m) to the matched backreef site
dist_to_br_m <- apply(dist_mat, 1, min)

## 3. Attach nearest backreef benthic info to PH sites ----

sites_benthic <- sites %>%
  mutate(
    br_site             = backreef_benthic$site[nn_idx],
    br_long             = backreef_benthic$long[nn_idx],
    br_lat              = backreef_benthic$lat[nn_idx],
    coral_cover_br      = backreef_benthic$coral_cover[nn_idx],
    macroalgae_cover_br = backreef_benthic$macroalgae_cover[nn_idx],
    ctb_cover_br        = backreef_benthic$ctb_cover[nn_idx],
    sand_cover_br       = backreef_benthic$sand_cover[nn_idx],
    millepora_cover_br  = backreef_benthic$millepora_cover[nn_idx],
    dist_to_br_m        = dist_to_br_m
  )

## 4. Inspect the mapping ----

sites_benthic %>%
  select(site.name, shore,
         br_site, dist_to_br_m,
         coral_cover_br, macroalgae_cover_br, ctb_cover_br,
         sand_cover_br, millepora_cover_br)

str(sites_benthic)

```

6. Fish functional groups: backreef fish, block-level
```{r}
# # Package ID: knb-lter-mcr.6.64 Cataloging System:https://pasta.edirepository.org.
# # Data set title: MCR LTER: Coral Reef: Long-term Population and Community Dynamics: Fishes, ongoing since 2005.
# # Data set creator:    - Moorea Coral Reef LTER 
# # Data set creator:  Andrew Brooks - Moorea Coral Reef LTER 
# # Metadata Provider:    - Moorea Coral Reef LTER 
# # Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu
# 
# #Read the main fish observations table (dt1 → fish_main)
# 
# options(HTTPUserAgent = "EDI_CodeGen")
# 
# ## ---------- Main fish table ----------
# 
# fish_main_url  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/6/64/ac2c7a859ce8595ec1339e8530b9ba50"
# fish_main_file <- tempfile()
# 
# try(
#   download.file(
#     fish_main_url,
#     fish_main_file,
#     method = "curl",
#     extra  = paste0(' -A "', getOption("HTTPUserAgent"), '"')
#   ),
#   silent = TRUE
# )
# 
# if (is.na(file.size(fish_main_file))) {
#   download.file(fish_main_url, fish_main_file, method = "auto")
# }
# 
# fish_main <- read.csv(
#   fish_main_file,
#   header = FALSE,
#   skip   = 1,
#   sep    = ",",
#   quote  = '"',
#   col.names = c(
#     "Year",
#     "Date",
#     "Start",
#     "End",
#     "Location",
#     "Site",
#     "Habitat",
#     "Depth",
#     "Transect",
#     "Swath",
#     "Taxonomy",
#     "Family",
#     "Count",
#     "Total_Length",
#     "Length_Anomaly",
#     "Biomass",
#     "Comment",
#     "Coarse_Trophic",
#     "Fine_Trophic",
#     "Cloud_Cover",
#     "Wind_Velocity",
#     "Sea_State",
#     "Swell",
#     "Visibility",
#     "Surge",
#     "Diver"
#   ),
#   stringsAsFactors = FALSE,
#   check.names = TRUE
# )
# 
# unlink(fish_main_file)
# 
# #Clean up types & sentinel values
# fish_main <- fish_main %>%
#   mutate(
#     Date  = as.Date(Date, format = "%Y-%m-%d"),
#     Start = as.factor(Start),
#     End   = as.factor(End),
# 
#     Location = as.factor(Location),
#     Site     = as.factor(Site),
#     Habitat  = as.factor(Habitat),
#     Transect = as.factor(Transect),
#     Swath    = as.factor(Swath),
# 
#     Taxonomy = ifelse(Taxonomy == "No fish observed", NA, Taxonomy),
#     Taxonomy = as.factor(Taxonomy),
#     Family   = as.factor(Family),
# 
#     # numeric fields
#     Depth         = suppressWarnings(as.numeric(Depth)),
#     Count         = suppressWarnings(as.numeric(Count)),
#     Total_Length  = suppressWarnings(as.numeric(Total_Length)),
#     Biomass       = suppressWarnings(as.numeric(Biomass)),
#     Cloud_Cover   = suppressWarnings(as.numeric(Cloud_Cover)),
#     Wind_Velocity = suppressWarnings(as.numeric(Wind_Velocity)),
#     Sea_State     = suppressWarnings(as.numeric(Sea_State)),
#     Swell         = suppressWarnings(as.numeric(Swell)),
#     Visibility    = suppressWarnings(as.numeric(Visibility)),
#     Surge         = suppressWarnings(as.numeric(Surge)),
# 
#     Length_Anomaly = ifelse(Length_Anomaly == "na", NA, Length_Anomaly),
#     Length_Anomaly = as.factor(Length_Anomaly),
# 
#     Comment       = ifelse(Comment == "NA", NA, Comment),
#     Comment       = as.factor(Comment),
# 
#     Coarse_Trophic = ifelse(Coarse_Trophic == "na", NA, Coarse_Trophic),
#     Coarse_Trophic = as.factor(Coarse_Trophic),
#     Fine_Trophic   = ifelse(Fine_Trophic == "na", NA, Fine_Trophic),
#     Fine_Trophic   = as.factor(Fine_Trophic),
# 
#     # sentinel → NA for lengths, biomass, visibility
#     Total_Length = ifelse(Total_Length == -1, NA, Total_Length),
#     Biomass      = ifelse(Biomass %in% c(-1), NA, Biomass),
#     Visibility   = ifelse(Visibility == -1, NA, Visibility),
# 
#     Diver = as.factor(Diver)
#   )


```

6.1 Filter fish data to backreef, set transect area, define trophic groups
```{r}

# Ensure Swath is numeric width (meters)
fish_main2 <- fish_main %>%
  mutate(
    Swath = as.numeric(as.character(Swath)),
    transect_length_m = 50,
    area_m2 = Swath * transect_length_m
  )

fish_backreef <- fish_main2 %>%
  filter(
    Habitat %in% c("Backreef"),
    Year >= 2020,
    !is.na(Biomass),
    !is.na(Count),
    area_m2 > 0
  ) %>%
  mutate(
    biomass_g = Biomass * Count,
    Site = gsub(" ", "_", Site),
    trophic_group = case_when(
      Fine_Trophic == "Herbivore/Detritivore" ~ "Herbivore/Detritivore",
      Fine_Trophic == "Corallivore" ~ "Corallivore",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(trophic_group))


```

6.2 Calculate fish biomass per m² per transect, then average per LTER backreef site
```{r}
# Biomass per m² at the TRANSECT level
fish_backreef_transect <- fish_backreef %>%
  group_by(Site, Year, Date, Transect, area_m2, trophic_group) %>%
  summarise(
    biomass_g = sum(biomass_g, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(biomass_g_m2 = biomass_g / area_m2) %>%
  ungroup()

## 1. Summarise biomass by site, year, and trophic group
fish_site_year_troph <- fish_backreef_transect %>%
  group_by(Site, Year, trophic_group) %>%
  summarise(
    mean_biomass_g_m2 = mean(biomass_g_m2, na.rm = TRUE),
    sd_biomass_g_m2   = sd(biomass_g_m2, na.rm = TRUE),
    n_transects       = dplyr::n(),
    se_biomass_g_m2   = sd_biomass_g_m2 / sqrt(n_transects)
  ) %>%
  ungroup()

## 2. Plot: stacked biomass by trophic group, per site, per year
trophic.colors <- c("coral", "darkgreen")

fishes <- ggplot(fish_site_year_troph,
       aes(x = Site,
           y = mean_biomass_g_m2,
           fill = trophic_group)) +
  geom_col() +
  scale_fill_manual (values=trophic.colors)+
  facet_wrap(~ Year, nrow = 1) +
  labs(
    x = "Site",
    y = "Mean fish biomass (g m²)",
    fill = "Trophic group",
    #title = "Back reef fish biomass by site, year, and trophic group"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
fishes

ggsave(
  filename = "Output/fishes_plot.png",
  plot = fishes,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)


# Now summarise to SITE level across years & transects
fish_backreef_site <- fish_backreef_transect %>%
  group_by(Site, trophic_group) %>%
  summarise(
    mean_biomass_g_m2 = mean(biomass_g_m2, na.rm = TRUE),
    sd_biomass_g_m2   = sd(biomass_g_m2, na.rm = TRUE),
    n_transects       = n(),
    .groups = "drop"
  ) %>%
  # Wide format: separate columns for herbivore / corallivore biomass
  pivot_wider(
    names_from = trophic_group,
    values_from = c(mean_biomass_g_m2, sd_biomass_g_m2, n_transects),
    names_glue = "fish_{trophic_group}_{.value}"
  )

fish_backreef_site


ggplot(fish_backreef_site, aes(x = Site, y = fish_Corallivore_mean_biomass_g_m2)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = fish_Corallivore_mean_biomass_g_m2 - fish_Corallivore_sd_biomass_g_m2, ymax = fish_Corallivore_mean_biomass_g_m2 + fish_Corallivore_sd_biomass_g_m2),
    width = 0.2,
    position = position_dodge(width = 0.7)
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Corallivore Fish Biomass (mean ± SD) by Site",
    y = "Biomass (g m⁻²)",
    x = "LTER Site"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#Fix the fish column names with /
fish_backreef_site <- fish_backreef_site %>%
  rename(
    fish_HerbDet_mean_biomass_g_m2 = `fish_Herbivore/Detritivore_mean_biomass_g_m2`,
    fish_HerbDet_sd_biomass_g_m2   = `fish_Herbivore/Detritivore_sd_biomass_g_m2`,
    fish_HerbDet_n_transects       = `fish_Herbivore/Detritivore_n_transects`
  )

ggplot(fish_backreef_site, aes(x = Site, y = fish_HerbDet_mean_biomass_g_m2)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = fish_HerbDet_mean_biomass_g_m2 - fish_HerbDet_sd_biomass_g_m2, ymax = fish_HerbDet_mean_biomass_g_m2 + fish_HerbDet_sd_biomass_g_m2),
    width = 0.2,
    position = position_dodge(width = 0.7)
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Herbivore/Detritivore Fish Biomass (mean ± SD) by Site",
    y = "Biomass (g m⁻²)",
    x = "LTER Site"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

6.3 Map fish metrics onto coral sites
```{r}
# sites_benthic must have: site.name, br_site
sites_temp_br <- sites_temp %>%
  left_join(
    sites_benthic %>% select(site.name, br_site),
    by = "site.name"
  )

str(sites_temp_br)

sites_temp_fish <- sites_temp_br %>%
  left_join(
    fish_backreef_site,
    by = c("br_site" = "Site")
  )

str(sites_temp_fish)

sites_fish <- sites_temp_fish %>%
  select(
    site.name, shore, br_site,
    fish_Corallivore_mean_biomass_g_m2,
    fish_HerbDet_mean_biomass_g_m2
  )

str(sites_fish)

```

7. Nutrients: Turbinaria N high-resolution, lagoon-wide
7.1 clean import + mapping of high-res Turbinaria CHN
```{r}
# Package ID: knb-lter-mcr.1047.1 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Turbinaria CHN from a spatially explicit sampling campaign in lagoons of Moorea, French Polynesia.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Thomas Adam - Moorea Coral Reef LTER 
# Data set creator:  Christian John - University of California, Santa Barbara 
# Data set creator:  Deron Burkepile - Moorea Coral Reef LTER 
# Metadata Provider:    - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu

options(HTTPUserAgent = "EDI_CodeGen")

## --------------------------------------------------
## TABLE 1: Site metadata (coords, habitat, distance)
## --------------------------------------------------
url_sites <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1047/1/45ed57442847daf0fb6a7dbccac56b53"
tmp_sites <- tempfile()
try(download.file(url_sites, tmp_sites,
                  method = "curl",
                  extra  = paste0(' -A "', getOption("HTTPUserAgent"), '"')))
if (is.na(file.size(tmp_sites))) {
  download.file(url_sites, tmp_sites, method = "auto")
}

turbinaria_sites <- read.csv(
  tmp_sites,
  header   = FALSE,
  skip     = 1,
  sep      = ",",
  quote    = '"',
  col.names = c(
    "Site",
    "Longitude",
    "Latitude",
    "Habitat",
    "Island_shore",
    "Distance_from_shore",
    "Distance_from_crest",
    "Watershed_area",
    "Cleared_area"
  ),
  check.names = TRUE,
  stringsAsFactors = FALSE
)

unlink(tmp_sites)

turbinaria_sites <- turbinaria_sites %>%
  mutate(
    Site                = factor(Site),
    Habitat             = factor(Habitat),
    Island_shore        = factor(Island_shore),
    Longitude           = as.numeric(Longitude),
    Latitude            = as.numeric(Latitude),
    Distance_from_shore = as.numeric(Distance_from_shore),
    Distance_from_crest = as.numeric(Distance_from_crest),
    Watershed_area      = as.numeric(Watershed_area),
    Cleared_area        = as.numeric(Cleared_area)
  )

str(turbinaria_sites)

## --------------------------------------------------
## TABLE 2: Full CHN measurements (Percent_N, C:N)
## --------------------------------------------------
url_chn <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1047/1/098ecd51f703743ef38e4db4967778cb"
tmp_chn <- tempfile()
try(download.file(url_chn, tmp_chn,
                  method = "curl",
                  extra  = paste0(' -A "', getOption("HTTPUserAgent"), '"')))
if (is.na(file.size(tmp_chn))) {
  download.file(url_chn, tmp_chn, method = "auto")
}

turbinaria_chn <- read.csv(
  tmp_chn,
  header   = FALSE,
  skip     = 1,
  sep      = ",",
  quote    = '"',
  col.names = c(
    "Site",
    "FieldSeason",
    "Date",
    "Year",
    "Month",
    "Day",
    "Weight_ug",
    "Percent_C",
    "Percent_H",
    "Percent_N",
    "C_to_N_ratio"
  ),
  check.names = TRUE,
  stringsAsFactors = FALSE
)

unlink(tmp_chn)

turbinaria_chn <- turbinaria_chn %>%
  mutate(
    Site        = factor(Site),
    FieldSeason = factor(FieldSeason),
    Date        = as.Date(Date, format = "%Y-%m-%d"),
    Year        = as.integer(Year),
    Month       = as.numeric(Month),
    Day         = as.numeric(Day),
    Weight_ug   = as.numeric(Weight_ug),
    Percent_C   = as.numeric(Percent_C),
    Percent_H   = as.numeric(Percent_H),
    Percent_N   = as.numeric(Percent_N),
    C_to_N_ratio = as.numeric(C_to_N_ratio)
  )

str(turbinaria_chn)


turbinaria_chn <- turbinaria_chn %>%
  filter(Year > 2016)

```

7.2 Summarize data and map onto the coral density site
```{r}
## --------------------------------------------------
## 1. Summarise Turbinaria %N by Turbinaria Site
## --------------------------------------------------

turb_N_site <- turbinaria_chn %>%
  group_by(Site) %>%
  summarise(
    turb_N_mean_pct = mean(Percent_N, na.rm = TRUE),
    turb_N_sd_pct   = sd(Percent_N,   na.rm = TRUE),
    turb_N_n_samples = sum(!is.na(Percent_N)),
    Year_min        = min(Year, na.rm = TRUE),
    Year_max        = max(Year, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # add coordinates + basic metadata
  left_join(
    turbinaria_sites %>%
      select(
        Site,
        turb_long         = Longitude,
        turb_lat          = Latitude,
        turb_Habitat      = Habitat,
        turb_Island_shore = Island_shore
      ),
    by = "Site"
  ) %>%
  # keep only sites with coordinates and N
  filter(!is.na(turb_long), !is.na(turb_lat), !is.na(turb_N_mean_pct))

## --------------------------------------------------
## 2. Coral sites from sites_wave
## --------------------------------------------------

coral_sites <- sites_wave %>%
  select(site.name, long, lat, shore) %>%
  distinct() %>%
  filter(!is.na(long), !is.na(lat))

## --------------------------------------------------
## 3. Haversine distance matrix (coral × Turbinaria)
## --------------------------------------------------

# Matrices of lon/lat (in degrees) for geosphere
coral_coords <- as.matrix(coral_sites[, c("long", "lat")])
turb_coords  <- as.matrix(turb_N_site[, c("turb_long", "turb_lat")])

# distHaversine can take vectors or matrices; use outer-style apply
# Build an n_coral × n_turb matrix of distances (m)
dist_mat <- sapply(
  1:nrow(turb_coords),
  function(j) distHaversine(coral_coords, turb_coords[j, ])
)

# dist_mat[i, j] = distance from coral_sites[i, ] to turb_N_site[j, ]

## --------------------------------------------------
## 4. For each coral site, find nearest Turbinaria site
## --------------------------------------------------

# index of nearest Turbinaria site per coral site
nn_index <- max.col(-dist_mat)  # argmin via argmax of negative

# minimum distance in meters
dist_min_m <- dist_mat[cbind(seq_len(nrow(coral_sites)), nn_index)]

## --------------------------------------------------
## 5. Build coral_turb lookup table
## --------------------------------------------------

coral_turb <- coral_sites %>%
  bind_cols(
    turb_N_site[nn_index, c("Site",
                            "turb_N_mean_pct",
                            "turb_N_sd_pct",
                            "turb_N_n_samples")] %>%
      rename(
        turb_site = Site
      )
  ) %>%
  mutate(
    dist_to_turb_m = dist_min_m
  )

# Inspect result
str(coral_turb)
head(coral_turb)

coral_turb$site.name <- factor(
  coral_turb$site.name,
  levels = c("S2", "S3", "S4", "S5", "S6", "S9", "S11", "S12", "S14", "S15", "S18" )
)


#  barplot: mean ± SD by site
nut_turb <- ggplot(coral_turb, aes(x = site.name, y = turb_N_mean_pct)) +
  geom_col(fill = "grey70", width = 0.6) +
  geom_errorbar(
    aes(ymin = turb_N_mean_pct - turb_N_sd_pct,
        ymax = turb_N_mean_pct + turb_N_sd_pct),
    width = 0.2
  ) +
  labs(
    #title = "Turbinaria tissue %N by coral survey site",
    x = "Site",
    y = "Turbinaria %N (mean ± SD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

nut_turb 

ggsave(
  filename = "Output/nutrients_turbinaria_plot.png",
  plot = nut_turb ,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)


```

8. fDOM Dissolved Organic Matter
8.1 clean fDOM import 
```{r}
# Package ID: knb-lter-mcr.1049.1 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Water Column fDOM in lagoons of Moorea, French Polynesia.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Craig Nelson - University of Hawaiʻi at Mānoa 
# Data set creator:  Sean Swift - University of Hawaiʻi at Mānoa 
# Metadata Provider:    - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu

## -------------------------------------------------------------
## fDOM data: site metadata and sample-level observations
## (non-overlapping names, no attach/detach, tidy-friendly)
## -------------------------------------------------------------

options(HTTPUserAgent = "EDI_CodeGen")

library(dplyr)

### 1. fDOM site metadata (coordinates, watershed, etc.) ----

inUrl_fdom_sites  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1049/1/45ed57442847daf0fb6a7dbccac56b53"
tmp_fdom_sites    <- tempfile()

try(download.file(
  inUrl_fdom_sites,
  tmp_fdom_sites,
  method = "curl",
  extra  = paste0(' -A "', getOption("HTTPUserAgent"), '"')
))
if (is.na(file.size(tmp_fdom_sites))) {
  download.file(inUrl_fdom_sites, tmp_fdom_sites, method = "auto")
}

fdom_sites_raw <- read.csv(
  tmp_fdom_sites,
  header    = FALSE,
  skip      = 1,
  sep       = ",",
  quote     = '"',
  col.names = c(
    "Site",
    "Longitude",
    "Latitude",
    "Habitat",
    "Island_shore",
    "Distance_from_shore",
    "Distance_from_crest",
    "Watershed_area",
    "Cleared_area"
  ),
  check.names = TRUE
)

unlink(tmp_fdom_sites)

## Basic type cleaning
fdom_sites_raw <- fdom_sites_raw %>%
  mutate(
    Site                = as.factor(Site),
    Habitat             = as.factor(Habitat),
    Island_shore        = as.factor(Island_shore),
    Longitude           = as.numeric(Longitude),
    Latitude            = as.numeric(Latitude),
    Distance_from_shore = as.numeric(Distance_from_shore),
    Distance_from_crest = as.numeric(Distance_from_crest),
    Watershed_area      = as.numeric(Watershed_area),
    Cleared_area        = as.numeric(Cleared_area)
  )

## Namespaced version to avoid clashes with other objects
fdom_sites <- fdom_sites_raw %>%
  rename(
    fdom_site          = Site,
    fdom_long          = Longitude,
    fdom_lat           = Latitude,
    fdom_habitat       = Habitat,
    fdom_shore_cat     = Island_shore,
    fdom_dist_shore_m  = Distance_from_shore,
    fdom_dist_crest_m  = Distance_from_crest,
    fdom_watershed_km2 = Watershed_area,
    fdom_cleared_km2   = Cleared_area
  )

str(fdom_sites)


### 2. fDOM sample-level data (HIX, FI, components, etc.) ----

inUrl_fdom_obs  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1049/1/c2fc28a7d3f0cec75aa163ac646d65be"
tmp_fdom_obs    <- tempfile()

try(download.file(
  inUrl_fdom_obs,
  tmp_fdom_obs,
  method = "curl",
  extra  = paste0(' -A "', getOption("HTTPUserAgent"), '"')
))
if (is.na(file.size(tmp_fdom_obs))) {
  download.file(inUrl_fdom_obs, tmp_fdom_obs, method = "auto")
}

fdom_obs_raw <- read.csv(
  tmp_fdom_obs,
  header    = FALSE,
  skip      = 1,
  sep       = ",",
  quote     = '"',
  col.names = c(
    "Site",
    "FieldSeason",
    "Date",
    "Year",
    "Month",
    "Day",
    "HIX",
    "BIX",
    "M.C",
    "FI",
    "Ultra.Violet.Humic.like",
    "Marine.Humic.like",
    "Visible.Humic.like",
    "Tryptophan.like",
    "Tyrosine.like",
    "Lignin.like"
  ),
  check.names = TRUE
)

unlink(tmp_fdom_obs)

## Basic type cleaning
fdom_obs_raw <- fdom_obs_raw %>%
  mutate(
    Site         = as.factor(Site),
    FieldSeason  = as.factor(FieldSeason),
    Date         = as.Date(Date, format = "%Y-%m-%d"),
    Year         = as.integer(Year),
    Month        = as.integer(Month),
    Day          = as.integer(Day),
    HIX          = as.numeric(HIX),
    BIX          = as.numeric(BIX),
    M.C          = as.numeric(M.C),
    FI           = as.numeric(FI),
    Ultra.Violet.Humic.like = as.numeric(Ultra.Violet.Humic.like),
    Marine.Humic.like       = as.numeric(Marine.Humic.like),
    Visible.Humic.like      = as.numeric(Visible.Humic.like),
    Tryptophan.like         = as.numeric(Tryptophan.like),
    Tyrosine.like           = as.numeric(Tyrosine.like),
    Lignin.like             = as.numeric(Lignin.like)
  )

## Namespaced version
fdom_obs <- fdom_obs_raw %>%
  rename(
    fdom_site          = Site,
    fdom_field_season  = FieldSeason,
    fdom_date          = Date,
    fdom_year          = Year,
    fdom_month         = Month,
    fdom_day           = Day,
    fdom_HIX           = HIX,
    fdom_BIX           = BIX,
    fdom_MC            = M.C,
    fdom_FI            = FI,
    fdom_UV_humic      = Ultra.Violet.Humic.like,
    fdom_marine_humic  = Marine.Humic.like,
    fdom_vis_humic     = Visible.Humic.like,
    fdom_tryptophan    = Tryptophan.like,
    fdom_tyrosine      = Tyrosine.like,
    fdom_lignin        = Lignin.like
  )

str(fdom_obs)

## Recalculate MC for issue in code of original paper.


fdom_obs <- fdom_obs %>%
  mutate(fdom_MC_corr = fdom_marine_humic/fdom_vis_humic)



```
8.2 Map fDOM to coral density sites
```{r}

## ------------------------------------------------------------
## 1. Summarise fDOM metrics per fDOM station
## ------------------------------------------------------------

fdom_site_summary <- fdom_obs %>%
  group_by(fdom_site) %>%
  summarise(
    fdom_HIX_mean          = mean(fdom_HIX, na.rm = TRUE),
    fdom_BIX_mean          = mean(fdom_BIX, na.rm = TRUE),
    fdom_MC_mean           = mean(fdom_MC,  na.rm = TRUE),
    fdom_MC_corr_mean      = mean(fdom_MC_corr,  na.rm = TRUE),
    fdom_FI_mean           = mean(fdom_FI,  na.rm = TRUE),
    fdom_UV_humic_mean     = mean(fdom_UV_humic,     na.rm = TRUE),
    fdom_marine_humic_mean = mean(fdom_marine_humic, na.rm = TRUE),
    fdom_vis_humic_mean    = mean(fdom_vis_humic,    na.rm = TRUE),
    fdom_tryptophan_mean   = mean(fdom_tryptophan,   na.rm = TRUE),
    fdom_tyrosine_mean     = mean(fdom_tyrosine,     na.rm = TRUE),
    fdom_lignin_mean       = mean(fdom_lignin,       na.rm = TRUE),
    fdom_n_samples         = dplyr::n(),
    .groups = "drop"
  )

## Attach coordinates + habitat info to those summaries
fdom_sites_full <- fdom_sites %>%
  inner_join(fdom_site_summary, by = "fdom_site")

## ------------------------------------------------------------
## 2. Extract unique coral-density sites with coords
## ------------------------------------------------------------

ph_sites <- sites %>%
  select(site.name, shore, long, lat) %>%
  distinct()

## ------------------------------------------------------------
## 3. Nearest-neighbor mapping: PH sites -> fDOM stations
## ------------------------------------------------------------

# All pairwise combinations
ph_fdom_pairs <- tidyr::crossing(
  ph_sites %>% rename(ph_site = site.name),
  fdom_sites_full
)

# Great-circle distance in meters
ph_fdom_pairs <- ph_fdom_pairs %>%
  mutate(
    dist_to_fdom_m = distHaversine(
      cbind(long, lat),
      cbind(fdom_long, fdom_lat)
    )
  )

# For each PH site, keep the closest fDOM station
sites_fdom <- ph_fdom_pairs %>%
  group_by(ph_site) %>%
  slice_min(dist_to_fdom_m, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(site.name = ph_site) %>%
  select(
    site.name, shore, long, lat,
    fdom_site, fdom_long, fdom_lat, fdom_habitat, fdom_shore_cat,
    dist_to_fdom_m,
    fdom_HIX_mean, fdom_BIX_mean, fdom_MC_mean, fdom_MC_corr_mean, fdom_FI_mean,
    fdom_UV_humic_mean, fdom_marine_humic_mean, fdom_vis_humic_mean,
    fdom_tryptophan_mean, fdom_tyrosine_mean, fdom_lignin_mean,
    fdom_n_samples
  )

## Quick check
print(sites_fdom)

# Select only the fDOM metrics (exclude n_samples)
fdom_metrics <- sites_fdom %>% 
  select(site.name,
         starts_with("fdom_")) %>%
  select(-fdom_site) %>%
  select(-fdom_long)%>%
  select(-fdom_lat)%>%
  select(-fdom_habitat)%>%
  select(-fdom_shore_cat)%>%
  select(-fdom_n_samples)

# Pivot longer
fdom_long <- fdom_metrics %>%
  pivot_longer(
    cols = -site.name,
    names_to = "metric",
    values_to = "value"
  )

# Plot faceted
ggplot(fdom_long, aes(x = site.name, y = value)) +
  geom_col(fill = "gray", width = 0.7) +
  facet_wrap(~ metric, scales = "free_y", ncol = 3) +
  labs(
    title = "fDOM metrics by site",
    x = "Site",
    y = "Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10, face = "bold")
  )

fdom_long2 <- sites_fdom %>% 
  select(site.name,shore,fdom_HIX_mean,fdom_BIX_mean,fdom_MC_corr_mean) %>%
  pivot_longer(
    cols = -c(site.name, shore),
    names_to = "metric",
    values_to = "value"
  )

fdom_long2$site.name <- factor(
  fdom_long2$site.name,
  levels = c("S2", "S3", "S4", "S5", "S6", "S9", "S11", "S12", "S14", "S15", "S18" )
)

fdom <- ggplot(fdom_long2, aes(x = site.name, y = value)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~ metric, scales = "free_y", ncol = 3) +
  labs(
   # title = "fDOM metrics by site",
    x = "Site",
    y = "Value",
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10, face = "bold")
  )
fdom 

ggsave(
  filename = "Output/fDOM_plot.png",
  plot = fdom  ,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)

```

9. Seawater Chemistry data (N,Si,P)
9.1 Import Seawater Chemistry data
```{r}
# Package ID: knb-lter-mcr.1048.1 Cataloging System:https://pasta.edirepository.org.
# Data set title: MCR LTER: Coral Reef: Water Column nutrient data in lagoons of Moorea, French Polynesia.
# Data set creator:    - Moorea Coral Reef LTER 
# Data set creator:  Thomas Adam - Moorea Coral Reef LTER 
# Data set creator:  Christian John - University of California, Santa Barbara 
# Data set creator:  Rebecca Vega Thurber - University of California, Santa Barbara 
# Data set creator:  Deron Burkepile - Moorea Coral Reef LTER 
# Metadata Provider:   Information Manager - Moorea Coral Reef LTER 
# Contact:    - Information Manager Moorea Coral Reef LTER  - mcrlter@msi.ucsb.edu

## ============================================================
## 1. IMPORT knb-lter-mcr.1048.1 (water-column nutrients)
##    - Site metadata  (table 1)
##    - Nutrient data  (table 2)
## ============================================================

options(HTTPUserAgent = "EDI_CodeGen")

## ---- 1a. Site metadata ----
nut_site_url  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1048/1/45ed57442847daf0fb6a7dbccac56b53"
nut_site_tmp  <- tempfile()

try(download.file(nut_site_url, nut_site_tmp, method = "curl",
                  extra = paste0(' -A "', getOption("HTTPUserAgent"), '"')))
if (is.na(file.size(nut_site_tmp))) {
  download.file(nut_site_url, nut_site_tmp, method = "auto")
}

nut_sites_raw <- read.csv(
  nut_site_tmp,
  header    = FALSE,
  skip      = 1,
  sep       = ",",
  quote     = '"',
  col.names = c(
    "Site",
    "Longitude",
    "Latitude",
    "Habitat",
    "Island_shore",
    "Distance_from_shore",
    "Distance_from_crest",
    "Watershed_area",
    "Cleared_area"
  ),
  check.names = TRUE,
  stringsAsFactors = FALSE
)

unlink(nut_site_tmp)

# Basic type cleanup & rename with non-overlapping names
nut_sites <- nut_sites_raw %>%
  mutate(
    Site              = factor(Site),
    Habitat           = factor(Habitat),
    Island_shore      = factor(Island_shore),
    Longitude         = as.numeric(Longitude),
    Latitude          = as.numeric(Latitude),
    Distance_from_shore  = as.numeric(Distance_from_shore),
    Distance_from_crest  = as.numeric(Distance_from_crest),
    Watershed_area       = as.numeric(Watershed_area),
    Cleared_area         = as.numeric(Cleared_area)
  ) %>%
  rename(
    nut_site_id             = Site,
    nut_long                = Longitude,
    nut_lat                 = Latitude,
    nut_habitat             = Habitat,
    nut_island_shore        = Island_shore,
    nut_dist_from_shore_m   = Distance_from_shore,
    nut_dist_from_crest_m   = Distance_from_crest,
    nut_watershed_area_km2  = Watershed_area,
    nut_cleared_area_km2    = Cleared_area
  )

## ---- 1b. Nutrient measurements ----
nut_data_url <- "https://pasta.lternet.edu/package/data/eml/knb-lter-mcr/1048/1/7c05cc39e4f71d6f49e0b0d592728e16"
nut_data_tmp <- tempfile()

try(download.file(nut_data_url, nut_data_tmp, method = "curl",
                  extra = paste0(' -A "', getOption("HTTPUserAgent"), '"')))
if (is.na(file.size(nut_data_tmp))) {
  download.file(nut_data_url, nut_data_tmp, method = "auto")
}

nut_samples_raw <- read.csv(
  nut_data_tmp,
  header    = FALSE,
  skip      = 1,
  sep       = ",",
  quote     = '"',
  col.names = c(
    "Site",
    "FieldSeason",
    "Date",
    "Year",
    "Month",
    "Day",
    "Phosphate",
    "Silicate",
    "Nitrite_plus_Nitrate",
    "Ammonia",
    "Notes"
  ),
  check.names = TRUE,
  stringsAsFactors = FALSE
)

unlink(nut_data_tmp)

# Parse dates & clean types, then rename columns
nut_samples <- nut_samples_raw %>%
  mutate(
    Site        = factor(Site),
    FieldSeason = factor(FieldSeason),
    Date        = as.Date(Date, format = "%Y-%m-%d"),
    Year        = as.integer(Year),
    Month       = as.integer(Month),
    Day         = as.integer(Day),
    Phosphate   = as.numeric(Phosphate),
    Silicate    = as.numeric(Silicate),
    Nitrite_plus_Nitrate = as.numeric(Nitrite_plus_Nitrate),
    Ammonia     = as.numeric(Ammonia),
    N_P         = as.numeric(Nitrite_plus_Nitrate)+ as.numeric(Ammonia)/as.numeric(Phosphate),
    Notes       = if (!is.null(Notes)) factor(Notes) else NULL
  ) %>%
  rename(
    nut_site_id    = Site,
    nut_season     = FieldSeason,
    nut_date       = Date,
    nut_PO4_uM     = Phosphate,
    nut_SiO4_uM    = Silicate,
    nut_NOx_uM     = Nitrite_plus_Nitrate,
    nut_NH4_uM     = Ammonia,
    nut_N_P        = N_P
  )

## ============================================================
## 2. SITE-LEVEL MEAN NUTRIENTS
## ============================================================

nut_site_means <- nut_samples %>%
  group_by(nut_site_id) %>%
  summarise(
    nut_PO4_mean_uM  = mean(nut_PO4_uM,  na.rm = TRUE),
    nut_SiO4_mean_uM = mean(nut_SiO4_uM, na.rm = TRUE),
    nut_NOx_mean_uM  = mean(nut_NOx_uM,  na.rm = TRUE),
    nut_NH4_mean_uM  = mean(nut_NH4_uM,  na.rm = TRUE),
    nut_N_P_mean = mean(nut_N_P,  na.rm = TRUE),
    nut_n_samples    = dplyr::n(),
    .groups = "drop"
  )

nut_sites_summary <- nut_site_means %>%
  left_join(
    nut_sites %>%
      select(
        nut_site_id,
        nut_long,
        nut_lat,
        nut_habitat,
        nut_island_shore,
        nut_dist_from_shore_m,
        nut_dist_from_crest_m,
        nut_watershed_area_km2,
        nut_cleared_area_km2
      ),
    by = "nut_site_id"
  )

```

9.2 clean and map data to coral sites
```{r}
## ============================================================
## 3. NEAREST–NEIGHBOUR MAP: NUTRIENT SITES → CORAL SITES
## ============================================================

# Coral sites from your existing env table
coral_sites <- sites %>%
  select(site.name, long, lat) %>%
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

coral_coords <- as.matrix(coral_sites[, c("long", "lat")])
nut_coords   <- as.matrix(nut_sites_summary[, c("nut_long", "nut_lat")])

# Index of nearest nutrient site for each coral site
nearest_idx_nut <- apply(coral_coords, 1, function(x) {
  dists_m <- distHaversine(x, nut_coords)
  which.min(dists_m)
})

# Actual distances (m)
nearest_dist_nut_m <- sapply(seq_len(nrow(coral_coords)), function(i) {
  distHaversine(coral_coords[i, ], nut_coords[nearest_idx_nut[i], ])
})

# Coral × nutrient table with mapped values
nut_by_coral <- coral_sites %>%
  mutate(
    nut_site_id_near        = nut_sites_summary$nut_site_id[nearest_idx_nut],
    nut_PO4_mean_uM         = nut_sites_summary$nut_PO4_mean_uM[nearest_idx_nut],
    nut_SiO4_mean_uM        = nut_sites_summary$nut_SiO4_mean_uM[nearest_idx_nut],
    nut_NOx_mean_uM         = nut_sites_summary$nut_NOx_mean_uM[nearest_idx_nut],
    nut_NH4_mean_uM         = nut_sites_summary$nut_NH4_mean_uM[nearest_idx_nut],
    nut_N_P_mean            = nut_sites_summary$nut_N_P_mean[nearest_idx_nut],
    nut_n_samples           = nut_sites_summary$nut_n_samples[nearest_idx_nut],
    nut_habitat_near        = nut_sites_summary$nut_habitat[nearest_idx_nut],
    nut_island_shore_near   = nut_sites_summary$nut_island_shore[nearest_idx_nut],
    dist_to_nut_site_m      = nearest_dist_nut_m
  )

nut_long <- nut_by_coral %>%
  select(site.name,
         nut_SiO4_mean_uM,
         nut_N_P_mean) %>%
  pivot_longer(
    cols = starts_with("nut_"),
    names_to = "nutrient",
    values_to = "value"
  ) %>%
  mutate(nutrient = recode(nutrient,
                           nut_SiO4_mean_uM = "SiO₄ (µM)",
                           nut_N_P_mean = "N:P"))

nut_long$site.name <- factor(
  nut_long$site.name,
  levels = c("S2", "S3", "S4", "S5", "S6", "S9", "S11", "S12", "S14", "S15", "S18" )
)

nut_water <- ggplot(nut_long, aes(x = site.name, y = value)) +
  geom_col(fill = "gray", width = 0.7) +
  facet_wrap(~ nutrient, scales = "free_y", ncol = 2) +
  labs(
    #title = "Water Column Nutrients by Coral Site",
    x = "Site",
    y = "Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1))
nut_water

ggsave(
  filename = "Output/nut_watercolumn_plot.png",
  plot = nut_water  ,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)

```



Recruitment
recruitment only at LTER1–2 (North), same for all PH sites → no spatial variation, so do not include it in X for this site-level JSDM. It can be used in interpretation only for temporal patterns that would assume equal recruitment around the island.

10. Build the predictor matrix X for boral

10.1 Join the environmental dataset
```{r}
#Start from temperature table (has long, lat, shore, temp metrics)
sites_env_full <- sites_temp %>%
  
  # ---- Add waves ----
  left_join(
    sites_wave %>%
      select(
        site.name,
        wave_P_mean = P_mean,
        wave_P_p95  = P_p95,
        wave_P_max  = P_max,
        wave_P_sd   = P_sd
      ),
    by = "site.name"
  ) %>%
  
  # ---- Add benthic cover (backreef) ----
  left_join(
    sites_benthic %>%
      select(
        site.name,
        br_site,
        br_long,
        br_lat,
        coral_cover_br,
        macroalgae_cover_br,
        ctb_cover_br,
        sand_cover_br,
        millepora_cover_br,
        dist_to_br_m
      ),
    by = "site.name"
  ) %>%
  
  # ---- Add fish biomass (backreef) ----
  left_join(
    sites_fish %>%
      select(
        site.name,
        fish_Corallivore_mean_biomass_g_m2,
        fish_HerbDet_mean_biomass_g_m2
      ),
    by = "site.name"
  ) %>%
  
  # ---- Add high spatial resolution nutrients (Turbinaria CHN) ----
  left_join(
    coral_turb %>%
      select(
        site.name,
        turb_N_mean_pct,
        turb_N_sd_pct,
        turb_N_n_samples
      ),
    by = "site.name"
  ) %>%

  # ---- Add fDOM ----
  left_join(
    sites_fdom %>%
      select(
        site.name,
        fdom_HIX_mean,
        fdom_BIX_mean,
        fdom_MC_corr_mean,
        fdom_FI_mean
      ),
    by = "site.name"
  ) %>%

  # ---- Add seawater chemistry ----
  left_join(
    nut_by_coral %>%
      select(
        site.name,
        nut_SiO4_mean_uM,
        nut_N_P_mean,
        nut_n_samples
      ),
    by = "site.name"
  )


# Have a look
glimpse(sites_env_full)

```

10.2 Decide which variables to keep as predictors
```{r}
# colnames(sites_env_full)
##add PCA here 

env_predictors_raw <- sites_env_full %>%
  select(
    site.name,
    max_temp_br,
    daily_range_br,
    wave_P_mean,
    coral_cover_br,
    #ctb_cover_br,
    #sand_cover_br,
    fish_Corallivore_mean_biomass_g_m2,
    #fish_HerbDet_mean_biomass_g_m2,
    turb_N_mean_pct,
    fdom_HIX_mean,
    fdom_BIX_mean,
    fdom_MC_corr_mean,
    nut_SiO4_mean_uM,
    nut_N_P_mean
  )

env_vars <- c(
    "max_temp_br",
    "daily_range_br",
    "wave_P_mean",
    "coral_cover_br",
    #"ctb_cover_br",
    #"sand_cover_br",
    "fish_Corallivore_mean_biomass_g_m2",
    #"fish_HerbDet_mean_biomass_g_m2",
    "turb_N_mean_pct",
    "fdom_HIX_mean",
    "fdom_BIX_mean",
    "fdom_MC_corr_mean",
    "nut_SiO4_mean_uM",
    "nut_N_P_mean"
  )
  

```

10.3 Scale the predictors for modeling
```{r}
# numeric-only matrix for scaling
X_env <- env_predictors_raw %>%
  select(-site.name) %>%
  scale() %>%
  as.data.frame()

# Add site names back for clarity
X_env <- cbind(site = env_predictors_raw$site.name, X_env)

# Look at scaled matrix
X_env

# Check for missing values (JSDMs cannot handle NAs without imputation)
colSums(is.na(X_env))

# Check distributions
summary(env_predictors_raw)

# Check correlations if worried about collinearity
cor(env_predictors_raw %>% select(-site.name), use = "pairwise.complete.obs")


# Select numeric predictors only
env_num <- env_predictors_raw %>%
  select(-site.name)

# Compute correlation matrix
cor_mat <- cor(env_num, use = "pairwise.complete.obs", method = "pearson")

cor_mat[upper.tri(cor_mat)] <- NA
cor_long <- melt(cor_mat, na.rm = TRUE)

predictor.corrs <- ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  
  # Add correlation values
  geom_text(aes(label = sprintf("%.2f", value)),
            color = "black", size = 3) +
  
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1,1), name = "Correlation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    x = "",
    y = "",
    title = "Correlation Matrix (Lower Triangle)"
  )

predictor.corrs

ggsave(
  filename = "Output/predictor_correlations_plot.png",
  plot = predictor.corrs,
  device = "png",
  width = 10,     # adjust width in inches
  height = 10,     # adjust height in inches
  units = "in"
)


```

SANITY CHECKS 
what predictor variable sites were chosen as closest to the coral density sites?
```{r}
plot_link_map <- function(df,
                          coral_lon, coral_lat,
                          pred_lon, pred_lat,
                          coral_label,
                          value_cols,
                          title_prefix = "") {
  
  df_long <- df %>%
    dplyr::select(all_of(c(coral_lon, coral_lat,
                           pred_lon, pred_lat,
                           coral_label,
                           value_cols))) %>%
    tidyr::pivot_longer(
      cols      = all_of(value_cols),
      names_to  = "variable",
      values_to = "value"
    )
  
  ggplot(df_long) +
    # line from coral site to predictor site
    geom_segment(
      aes(
        x    = .data[[coral_lon]],
        y    = .data[[coral_lat]],
        xend = .data[[pred_lon]],
        yend = .data[[pred_lat]]
      ),
      colour = "grey70",
      alpha  = 0.7
    ) +
    # predictor station (always black)
    geom_point(
      aes(
        x = .data[[pred_lon]],
        y = .data[[pred_lat]]
      ),
      size  = 2.3,
      shape = 21,
      fill  = "black",
      colour = "black"
    ) +
    # coral site (white with black border)
    geom_point(
      aes(
        x = .data[[coral_lon]],
        y = .data[[coral_lat]]
      ),
      shape  = 21,
      fill   = "white",
      size   = 2.8,
      colour = "black"
    ) +
    # coral site labels
    geom_text(
      aes(
        x     = .data[[coral_lon]],
        y     = .data[[coral_lat]],
        label = .data[[coral_label]]
      ),
      vjust = -0.7,
      size  = 2.8
    ) +
    coord_equal() +  # keep true geography
    facet_wrap(~ variable, ncol = 2) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid        = element_blank(),
      strip.background  = element_rect(fill = "grey95", colour = NA),
      axis.title.x      = element_text(margin = margin(t = 5)),
      axis.title.y      = element_text(margin = margin(r = 5)),
      legend.position   = "none"     # remove color legend entirely
    ) +
    labs(
      x     = "Longitude",
      y     = "Latitude",
      title = title_prefix
    )
}


```

fDOM
```{r}
fdom_value_cols <- c(
  "fdom_HIX_mean"
)

fdom_spatial <- plot_link_map(
  df          = sites_fdom,
  coral_lon   = "long",
  coral_lat   = "lat",
  pred_lon    = "fdom_long",
  pred_lat    = "fdom_lat",
  coral_label = "site.name",
  value_cols  = fdom_value_cols,
  title_prefix = "fDOM metrics – coral survey sites and nearest fDOM stations"
)
fdom_spatial

ggsave(
  filename = "Output/fdom_spatial_plot.png",
  plot = fdom_spatial,
  device = "png",
  width = 10,     # adjust width in inches
  height = 10,     # adjust height in inches
  units = "in"
)


```
Turbinaria high spatial resolation
```{r}
## From the nutrient spatial metadata (1048 dt1)
nut_spatial_coords <- turbinaria_sites %>%      # <-- whatever you named dt1
  dplyr::select(
    nut_site_id = Site,
    nut_long    = Longitude,
    nut_lat     = Latitude
  )

## Attach coordinates of the nearest nutrient site to each coral site
nut_by_coral_loc <- nut_by_coral %>%
  dplyr::left_join(
    nut_spatial_coords,
    by = c("nut_site_id_near" = "nut_site_id")
  )


nut_value_cols <- c(
  "nut_SiO4_mean_uM"
)

nut_water_spatial <- plot_link_map(
  df          = nut_by_coral_loc,
  coral_lon   = "long",
  coral_lat   = "lat",
  pred_lon    = "nut_long",
  pred_lat    = "nut_lat",
  coral_label = "site.name",
  value_cols  = nut_value_cols,
  title_prefix = "Nutrient concentrations – coral sites and nearest nutrient stations"
)
nut_water_spatial

ggsave(
  filename = "Output/nut_water_spatial_plot.png",
  plot = nut_water_spatial,
  device = "png",
  width = 10,     # adjust width in inches
  height = 10,     # adjust height in inches
  units = "in"
)


```
Benthic Data
```{r}
benthic_value_cols <- c(
  "coral_cover_br"
)

benthic_spatial <- plot_link_map(
  df          = sites_benthic,
  coral_lon   = "long",
  coral_lat   = "lat",
  pred_lon    = "br_long",
  pred_lat    = "br_lat",
  coral_label = "site.name",
  value_cols  = benthic_value_cols,
  title_prefix = "Back-reef benthic cover – coral sites and nearest LTER benthic station"
)
benthic_spatial

ggsave(
  filename = "Output/benthic_spatial_plot.png",
  plot = benthic_spatial,
  device = "png",
  width = 10,     # adjust width in inches
  height = 10,     # adjust height in inches
  units = "in"
)

```

Temperature Data
```{r}
temp_value_cols <- c(
  "max_temp_br"
)

temp_spatial <- plot_link_map(
  df          = sites_temp,
  coral_lon   = "long",
  coral_lat   = "lat",
  pred_lon    = "temp_long",
  pred_lat    = "temp_lat",
  coral_label = "site.name",
  value_cols  = temp_value_cols,
  title_prefix = "Temperature – coral sites and nearest LTER thermister array"
)
temp_spatial

ggsave(
  filename = "Output/temp_spatial_plot.png",
  plot = temp_spatial,
  device = "png",
  width = 10,     # adjust width in inches
  height = 10,     # adjust height in inches
  units = "in"
)

```


11. Fit the boral JSDM
11.1 Align Y and X (by site names)
```{r}
# Extract site order from Y
Y_sites <- rownames(Y)

# Reorder X_env to match Y
X_env_aligned <- X_env %>%
  arrange(match(site, Y_sites))

# Verify alignment
stopifnot(all(rownames(Y) == X_env_aligned$site))

# Build predictor matrix X
X <- X_env_aligned %>%
  select(-site) %>%
  as.matrix()

```


11.2 Run the model
```{r}
set.seed(123)

boral_fit <- boral(
  y = Y,
  X = X,
  family = "normal",
  lv.control = list(num.lv = 2),
  row.eff = "none",
  save.model = TRUE,
  mcmc.control = list(
    n.burnin = 20000,
    n.iteration = 40000,
    n.thin = 20
  )
)
```

12. Model output
12.1 Model output: coefficients
```{r}

names(boral_fit)

#Environmental coefficients (predictor → taxon responses)
X_eff <- boral_fit$X.coefs.median
X_eff

X_long <- melt(X_eff, varnames = c("response", "predictor"), value.name = "beta")

ggplot(X_long, aes(predictor, response, fill = beta)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Environmental predictor",
    y = "Species × size class",
    title = "boral: posterior median regression coefficients"
  )


```
12.2 Residual correlations
```{r}
res_cor <- get.residual.cor(boral_fit)

res_long <- melt(res_cor, varnames = c("response1", "response2"), 
                 value.name = "resid_cor") %>%
  filter(!is.na(response1))

ggplot(res_long, aes(response1, response2, fill = resid_cor)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166AC", mid = "gray", high = "#B2182B",
    midpoint = 0, limits = c(-1,1)
  ) +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Residual correlation matrix (boral)")

```

12.3 Plot taxa ordination
```{r}
## Extract LV loadings from boral
loadings <- boral_fit$lv.coefs.median
L <- as.data.frame(loadings)
L$taxon <- rownames(L)

## Extract species name (before first underscore)
L$species <- sub("_.*", "", L$taxon)

## Define your custom species color palette
species_cols <- c(
  "P. tuahiniensis" = "#D55E00",
  "P. meandrina"    = "#CC79A7",
  "P. verrucosa"    = "#E69F00",
  "P. grandis"      = "#0072B2",
  "P. effusa"       = "#009E73",
  "P. acuta"        = "#e63946"
)

## Plot LV1 vs LV2 with species colors
ggplot(L, aes(theta1, theta2, label = taxon, color = species)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.6, size = 3) +
  scale_color_manual(values = species_cols) +
  theme_bw() +
  labs(
    title = "Taxon Loadings in LV Space (boral)",
    x = "LV1 (theta1)",
    y = "LV2 (theta2)",
    color = "Species"
  )

#zoom in by excluding visualization of P. acuta
## 2. Extract species
L$species <- sub("_.*", "", L$taxon)

## 3. Remove P. acuta rows
L_no_acuta <- L %>% 
  filter(species != "P. acuta")

## 4. Custom colors
species_cols <- c(
  "P. tuahiniensis" = "#D55E00",
  "P. meandrina"    = "#CC79A7",
  "P. verrucosa"    = "#E69F00",
  "P. grandis"      = "#0072B2",
  "P. effusa"       = "#009E73"
  # P. acuta intentionally omitted
)

## 5. LV scatter plot without P. acuta
ggplot(L_no_acuta, aes(theta1, theta2, label = taxon, color = species)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.6, size = 3) +
  scale_color_manual(values = species_cols) +
  theme_bw() +
  labs(
    title = "Taxon Loadings in LV Space (P. acuta removed)",
    x = "LV1 (theta1)",
    y = "LV2 (theta2)",
    color = "Species"
  )


```


12.4 Plot site ordination
```{r}
# Extract site scores (latent variables) – boral stores these as lv.median
site_lv <- as.data.frame(boral_fit$lv.median)

# Add site names (should match rownames of Y)
site_lv$site <- rownames(Y)

str(site_lv)
# Expect columns: LV1, LV2 (and maybe LV3... if num.lv > 2)

# Example using sites_env_full; adjust object name if needed
site_lv <- site_lv %>%
  left_join(
    sites_env_full %>% select(site.name, shore),
    by = c("site" = "site.name")
  )

ggplot(site_lv, aes(x = lv1, y = lv2, label = site, color = shore)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.05, size = 3) +
  theme_minimal() +
  labs(
    title = "Site ordination in latent variable space (boral)",
    x = "Latent Variable 1 (LV1)",
    y = "Latent Variable 2 (LV2)",
    color = "Shore"
  )

site_lv_env <- site_lv %>%
  left_join(X_env, by = "site")   # both have 'site' = S11, S12, ...

ggplot(site_lv_env, aes(x = lv1, y = lv2)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey85") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey85") +
  geom_point(aes(color = wave_P_mean, shape = shore), size = 3) +
  geom_text(aes(label = site), nudge_y = 0.05, size = 3) +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Site ordination with max wave power",
    x = "Latent Variable 1 (LV1)",
    y = "Latent Variable 2 (LV2)",
    color = "Average Wave Power\n(z-score)",
    shape = "Shore"
  )



## --- 3. Choose environmental variables to show as arrows -----------------
env_vars #listed above in model

# Matrix of LV scores
lv_mat   <- as.matrix(site_lv_env[, c("lv1", "lv2")])

# Matrix of environmental predictors (must be numeric)
env_mat  <- as.matrix(site_lv_env[, env_vars])

## --- 4. Compute environmental arrow directions (correlations) -----------
# For each env variable, correlate with LV1 and LV2
env_arrows <- lapply(seq_along(env_vars), function(j) {
  vname <- env_vars[j]
  x     <- env_mat[, j]

  c(
    var   = vname,
    cor_LV1 = cor(x, lv_mat[, 1], use = "pairwise.complete.obs"),
    cor_LV2 = cor(x, lv_mat[, 2], use = "pairwise.complete.obs")
  )
})

env_arrows <- do.call(rbind, env_arrows) %>%
  as.data.frame() %>%
  mutate(
    cor_LV1 = as.numeric(cor_LV1),
    cor_LV2 = as.numeric(cor_LV2)
  )

# Scale arrow lengths so they fit nicely in the LV space
arrow_scale <- 1.5
env_arrows <- env_arrows %>%
  mutate(
    xend = cor_LV1 * arrow_scale,
    yend = cor_LV2 * arrow_scale
  )

## --- 5. Site-ordination biplot: shore ellipses + env arrows -------------
ggplot() +
  # Ellipses by shore (site groups)
  stat_ellipse(
    data = site_lv_env,
    aes(x = lv1, y = lv2, color = shore, group = shore),
    type  = "norm",
    level = 0.68,
    linewidth = 0.6,
    alpha = 0.4
  ) +
  # Sites
  geom_point(
    data = site_lv_env,
    aes(x = lv1, y = lv2, color = shore),
    size = 3
  ) +
  ggrepel::geom_text_repel(
    data = site_lv_env,
    aes(x = lv1, y = lv2, label = site, color = shore),
    size = 3,
    max.overlaps = 50,
    show.legend = FALSE
  ) +
  # Axes through origin
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  # Environmental arrows (from origin)
  geom_segment(
    data = env_arrows,
    aes(
      x = 0, y = 0,
      xend = xend,
      yend = yend
    ),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "grey20",
    linewidth = 0.4
  ) +
  # Labels for environmental variables
  ggrepel::geom_text_repel(
    data = env_arrows,
    aes(
      x = xend,
      y = yend,
      label = var
    ),
    size = 3,
    color = "black",
    max.overlaps = 100,
    segment.color = NA
  ) +
  labs(
    x = "LV1 (site scores)",
    y = "LV2 (site scores)",
    color = "Shore",
    title = "Site ordination in LV space with environmental gradients",
    subtitle = "Ellipses = shore groups; arrows = env gradients (correlation with LV1/LV2)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right"
  )


```



13. Variance partitioning 
13.1 identify which predictors matter most overall
```{r}
#coefs <- boral_fit$jags.model$BUGSoutput$sims.list


## Columns you used as predictors in boral (order matters!)
env_cols <- env_vars

## Match X rows to Y rows by site name
X_env_ordered <- X_env %>%
  filter(site %in% rownames(Y)) %>%        # keep the 11 PH sites
  arrange(match(site, rownames(Y)))        # match the order of Y

stopifnot(all(X_env_ordered$site == rownames(Y)))

## X matrix used in the model
X_mat <- as.matrix(X_env_ordered[, env_cols])
rownames(X_mat) <- X_env_ordered$site

n_site    <- nrow(Y)
n_species <- ncol(Y)
p_X       <- ncol(X_mat)

species_names <- colnames(Y)
pred_names    <- colnames(X_mat)

## ==========================================================
## 2. Extract posterior β (X.coefs) from boral_fit
## ==========================================================

coefs <- boral_fit$jags.model$BUGSoutput$sims.list

if (!"X.coefs" %in% names(coefs)) {
  stop("X.coefs not found in boral_fit$jagsoutput$BUGSoutput$sims.list")
}

beta_array_full <- coefs$X.coefs      # MCMC samples of X coefficients
dim_beta <- dim(beta_array_full)      # usually: n_sims × something × something

n_sims <- dim_beta[1]

## Figure out which dimension is species, which is coefficients
if (dim_beta[2] == n_species) {
  # iter × species × coef
  beta_array_full <- aperm(beta_array_full, c(1, 3, 2))  # → iter × coef × species
} else if (dim_beta[3] == n_species) {
  # iter × coef × species → already in desired form
  # do nothing
} else {
  stop("Cannot match dimensions of X.coefs to n_species. ",
       "dim(X.coefs) = ", paste(dim_beta, collapse = " x "),
       ", n_species = ", n_species)
}

n_coef_full <- dim(beta_array_full)[2]

## Expect: intercept + p_X predictors  OR just p_X predictors
if (n_coef_full == p_X + 1) {
  message("Detected intercept in X.coefs; dropping the first column.")
  beta_array <- beta_array_full[, -1, , drop = FALSE]  # iter × p_X × species
} else if (n_coef_full == p_X) {
  message("No separate intercept column detected in X.coefs; using all coefficients as predictors.")
  beta_array <- beta_array_full                          # iter × p_X × species
} else {
  stop("Number of X coefficients (", n_coef_full,
       ") does not match number of predictors (", p_X,
       " or ", p_X + 1, "). Check that X_env and boral model use the same predictors and order.")
}

## Posterior mean β (predictor × species)
beta_mean <- apply(beta_array, c(2, 3), mean)   # p_X × n_species
rownames(beta_mean) <- pred_names
colnames(beta_mean) <- species_names

## ==========================================================
## 3. Variance partitioning by species
## ==========================================================
## For each species j:
##   eta_total_ij = sum_k X_ik * β_kj   (over predictors k)
##   eta_k_ij     = X_ik * β_kj         (effect of predictor k alone)
##   var_kj       = Var_i(eta_k_·j)
## We normalise var_kj so proportions per species sum to 1.

varpart_mat <- matrix(NA_real_, nrow = p_X, ncol = n_species,
                      dimnames = list(pred_names, species_names))

for (j in seq_len(n_species)) {
  # fitted linear predictor across sites for species j
  eta_total_j <- as.numeric(X_mat %*% beta_mean[, j])  # n_site vector
  
  # raw variance contribution of each predictor k to species j
  var_k_j <- numeric(p_X)
  
  for (k in seq_len(p_X)) {
    eta_k_j <- X_mat[, k] * beta_mean[k, j]  # effect of this predictor only
    var_k_j[k] <- stats::var(eta_k_j)
  }
  
  # normalise so contributions sum to 1 for each species
  if (sum(var_k_j) > 0) {
    varpart_mat[, j] <- var_k_j / sum(var_k_j)
  } else {
    varpart_mat[, j] <- NA_real_
  }
}

## ==========================================================
## 4. Summarise across species (overall predictor importance)
## ==========================================================

overall_importance <- rowMeans(varpart_mat, na.rm = TRUE)
overall_importance <- sort(overall_importance, decreasing = TRUE)

overall_importance

## ==========================================================
## 5. Plot variance partitioning
## ==========================================================

## Heatmap: predictors × species
varpart_long <- melt(varpart_mat, varnames = c("Predictor", "Species"),
                     value.name = "prop_var")

ggplot(varpart_long, aes(x = Species, y = Predictor, fill = prop_var)) +
  geom_tile(color = "grey80") +
  scale_fill_gradient2(
    low = "white", mid = "#FDB863", high = "#B2182B",
    midpoint = 0.25, na.value = "grey90",
    name = "Proportion\nof variance"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Approximate variance partitioning by species",
    x = "Coral species × size class",
    y = "Predictor"
  )

## Barplot: overall importance across species
overall_df <- data.frame(
  Predictor = names(overall_importance),
  MeanProp  = as.numeric(overall_importance)
)

ggplot(overall_df, aes(x = reorder(Predictor, MeanProp), y = MeanProp)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 11) +
  labs(
    title = "Average proportional contribution to linear predictor variance",
    x = "Predictor",
    y = "Mean proportion across species"
  )

```

13.X Environmental predictors 
Single/pairwise predictor effects per site
```{r}
## 1. Extract β (environmental coefficients) and X (predictor matrix)
B  <- boral_fit$X.coefs.median          # species × predictors (s × p)
BX <- as.matrix(boral_fit$X)            # sites × predictors   (n × p)

## 2. Compute environmental influence: E = X β  → n_sites × n_taxa
# (B is s × p, so we need t(B) = p × s)
E_env <- BX %*% t(B)                    # 11 × 24  (sites × taxa)

## 3. Put into a data.frame
E_env_df <- as.data.frame(E_env)

## 4. Recover site names from Y / boral_fit$y
site_names <- rownames(boral_fit$y)
if (is.null(site_names)) {
  site_names <- rownames(Y)  # fallback if you still have Y in your workspace
}

E_env_df$site.name <- site_names

## 5. Long format
E_long <- E_env_df %>%
  relocate(site.name) %>%
  pivot_longer(
    cols      = -site.name,
    names_to  = "taxon",
    values_to = "env_influence"
  )

## 6. Heatmap
ggplot(E_long, aes(x = site.name, y = taxon, fill = env_influence)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0
  )
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Environmental contribution (Xβ) by site and taxon",
    x = "Site",
    y = "Species × size class",
    fill = "Xβ"
  )

```
decomposition by predictor
```{r}

## ---------------------------------------------------------
## 0. Extract matrices
## ---------------------------------------------------------

# β coefficients (taxon × predictor) → transpose to predictor × taxon
B_full <- boral_fit$X.coefs.median
if (is.null(B_full)) stop("X.coefs.median not found in boral_fit.")
B_full <- t(B_full)   # predictors × taxa

# X used in the model
X_full <- as.data.frame(boral_fit$X)

# Use boral's site names
site_names_true <- rownames(boral_fit$lv.median)
rownames(X_full) <- site_names_true

## ---------------------------------------------------------
## 1. Align predictors between X and B
## ---------------------------------------------------------

common_pred <- intersect(colnames(X_full), rownames(B_full))
if (length(common_pred) == 0)
  stop("No matching predictor names between X and X.coefs.median")

X <- as.matrix(X_full[, common_pred, drop = FALSE])      # n_site × n_pred
B <- B_full[common_pred, , drop = FALSE]                 # n_pred × n_taxa

n_site <- nrow(X)
n_pred <- ncol(X)
n_taxa <- ncol(B)

site_names <- rownames(X)
taxa_names <- colnames(B)
pred_names <- colnames(X)

## ---------------------------------------------------------
## 2. Build contrib_arr[site, taxon, predictor] = X_ik * β_kj
## ---------------------------------------------------------
contrib_arr <- array(
  NA_real_,
  dim = c(n_site, n_taxa, n_pred),
  dimnames = list(
    site      = site_names,
    taxon     = taxa_names,
    predictor = pred_names
  )
)

for (k in seq_len(n_pred)) {
  contrib_arr[ , , k] <- outer(X[ , k], B[k, ])  # site × taxon
}

## ---------------------------------------------------------
## 3. Collapse across taxa → mean |Xβ| per site × predictor
## ---------------------------------------------------------
site_pred_influence <- apply(abs(contrib_arr), c(1, 3), mean, na.rm = TRUE)
# n_site × n_pred

## ---------------------------------------------------------
## 4. Normalize per site → fractions that sum to 1
## ---------------------------------------------------------
site_pred_fraction <- site_pred_influence / rowSums(site_pred_influence)

## ---------------------------------------------------------
## 5. Cluster predictors (columns) AND sites (rows)
## ---------------------------------------------------------

# # Cluster predictors
# pred_dist  <- dist(t(site_pred_fraction))          # predictors
# pred_hcl   <- hclust(pred_dist, method = "complete")
# pred_order <- pred_hcl$labels[pred_hcl$order]
# 
# # Cluster sites
# site_dist  <- dist(site_pred_fraction)             # sites
# site_hcl   <- hclust(site_dist, method = "complete")
# site_order <- site_hcl$labels[site_hcl$order]

## ---------------------------------------------------------
## 6. Long format with clustered factor levels
## ---------------------------------------------------------
site_pred_frac_long <- as.data.frame(as.table(site_pred_fraction))
colnames(site_pred_frac_long) <- c("site", "predictor", "fraction")

site_pred_frac_long$predictor <- factor(
  site_pred_frac_long$predictor
)

site_pred_frac_long$site <- factor(
  site_pred_frac_long$site
)

## ---------------------------------------------------------
## 7. Heatmap: clustered by both site and predictor
## ---------------------------------------------------------

site_pred_frac_long$site <- factor(
  site_pred_frac_long$site,
  levels = c("S2", "S3", "S4", "S5", "S6", "S9", "S11", "S12", "S14", "S15", "S18" )
)

influence <- ggplot(site_pred_frac_long, aes(x = predictor, y = site, fill = fraction)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(
    option = "D",
    name   = "Fraction of\nXβ contribution"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid  = element_blank()
  ) +
  labs(
    x = "Predictor",
    y = "Site",
    title = "Environmental predictor influence fractions"
  )

influence

ggsave(
  filename = "Output/predictor_influence_plot.png",
  plot = influence,
  device = "png",
  width = 8,     # adjust width in inches
  height = 4,     # adjust height in inches
  units = "in"
)

#######
### WAIC, need to add it to the model to check the sanity 

```


