---
title: "Untitled"
author: "HM Putnam"
date: "2024-07-26"
output: html_document
---
load libraries
```{r}
library(tidyverse)
library(ggmap)
#library(ggsn)
library(gridExtra)
library(cowplot)
library(dplyr)
library(tidyr)
```


load master list to count samples per site
```{r}
all.data <- read.csv("Data/POC_ID_RA_list.csv")%>%
  filter(Timing=="Pre_bleach")

sample.counts <- all.data %>%
  group_by(Site) %>%
  summarise(total = n()) 
sample.counts

sample.transect.counts <- all.data %>%
  group_by(Site, Transect) %>%
  summarise(total = n()) 
sample.transect.counts 

```

load size data
```{r}
size.data <- read.csv("Data/POC_RA_Size.csv")

# # Set a seed for reproducibility (optional)
# set.seed(123)
# 
# # Fill Diameter.1 with random values from 0.1 to 1000
# size.data$Diameter.1 <- runif(nrow(size.data), min = 0.1, max = 1000)
# 
# # Function to generate a value within 2 units of a given value
# generate_within_two <- function(value) {
#   min_val <- max(0.1, value - 2)  # Ensure the value does not go below 0.1
#   max_val <- value + 2
#   runif(1, min = min_val, max = max_val)
# }
# 
# # Fill Diameter.2 and Diameter.3
# size.data$Diameter.2 <- sapply(size.data$Diameter.1, generate_within_two)
# size.data$Diameter.3 <- sapply(size.data$Diameter.1, generate_within_two)


# Calculate the average of the Diameter columns
size.data <- size.data %>%
  mutate(Average.Diameter.cm = rowMeans(select(., Diameter.1, Diameter.2, Diameter.3), na.rm = TRUE))


```

```{r}
#code for plotting only the haplotype and not by species on the graph 
sample.info <- read.csv("Data/sample.info.csv")

data_h <- read.csv("Data/Haplotype_data.csv", na.strings = "NA")

data_h <- left_join(data_h, sample.info, by="Sample.ID")


# set the colors for plotting
 cols.haplotype <- cbind.data.frame(Haplotype = factor(c("Haplotype 10",
                                            "Haplotype 1a",
                                            "Haplotype 3",
                                             "P. grandis",
                                             "Haplotype 2",
                                             "Haplotype 11",
                                           "Haplotype 5a",
                                             "Haplotype 8a")),
                          Color = c("#D55E00",
                                  "#0072B2",
                                   "#E69F00",
                                   "#56B4E9",
                                   "lightgreen",
                                   "#009E73",
                                   "#e63946",
                                   "#CC79A7"))
                                   
# color_mapping <- setNames(cols.haplotype$Color, cols.haplotype$Haplotype)

                                   
color_mapping_h <- setNames(cols.haplotype$Color, cols.haplotype$Haplotype)
```

load abundance data
```{r}
sample.info <- read.csv("Data/sample.info.csv")

data <- read.csv("Data/Haplotype_data.csv", na.strings = "NA")

data <- left_join(data, sample.info, by="Sample.ID")


# set the colors for plotting
# cols.haplotype <- cbind.data.frame(Haplotype = factor(c("Haplotype 10",
#                                             "Haplotype 1a",
#                                             "Haplotype 3",
#                                             #"P. grandis",
#                                             "Haplotype 2",
#                                             "Haplotype 11",
#                                             "Haplotype 5a",
#                                             "Haplotype 8a")),
#                          Color = c("#D55E00",
#                                    "#0072B2",
#                                    "#E69F00",
#                                    #"#56B4E9",
#                                    "#009E73",
#                                    "#009E73",
#                                    "#e63946",
#                                    "#CC79A7"))
#                                    
# color_mapping <- setNames(cols.haplotype$Color, cols.haplotype$Haplotype)


cols.species <- cbind.data.frame(Species = factor(c("P. tuahiniensis",
                                            "P. meandrinaP. grandis",
                                            "P. verrucosa",
                                            "P. grandis",
                                            "P. effusa",
                                            "P. acuta",
                                            "P. meandrina")),
                         Color = c("#D55E00",
                                   "#0072B2",
                                   "#E69F00",
                                   "#56B4E9",
                                   "#009E73",
                                   "#e63946",
                                   "#CC79A7"))
                                   
color_mapping <- setNames(cols.species$Color, cols.species$Species)



```
Join id and size and set size categories for ploting with haplotype
```{r}

colnames(data_h)
colnames(size.data)



data_h$Sample.ID <- as.integer(data_h$Sample.ID)


data_h <- left_join(data_h, size.data, by = "Sample.ID")

data_h <- data_h %>%
  select(Sample.ID, Site.x, Transect.x,Haplotype, Average.Diameter.cm)

colnames(data_h) <- c("Sample.ID", "Site", "Transect", "Haplotype", "Average.Diameter.cm")

# Define the break points for the categories
breaks <- c(0,1, 2,5, 14, Inf)  # Adjust these values as needed
labels <- c("0-1cm", "1-2cm", ">2cm-5cm", ">5cm-14cm", ">14cm")  # Category labels

# Create a new column for LifeStage based on Average.Diameter.cm
data_h$LifeStage <- cut(data_h$Average.Diameter.cm, 
                      breaks = breaks, 
                      labels = labels, 
                      right = FALSE)  # Use right=FALSE to make intervals [a, b)

#removing the raw with NA for sizes because we don't have the pics to extract the data
data_clean <- data_h %>%
  filter(if_all(everything(), ~ !is.na(.)))



```
Join id and size and set size categories
```{r}

colnames(data)
colnames(size.data)



data$Sample.ID <- as.integer(data$Sample.ID)


data <- left_join(data, size.data, by = "Sample.ID")

data <- data %>%
  select(Sample.ID, Site.x, Transect.x,Species, Average.Diameter.cm)

colnames(data) <- c("Sample.ID", "Site", "Transect", "Species", "Average.Diameter.cm")

# Define the break points for the categories
breaks <- c(0,2, 5, 14, Inf)  # Adjust these values as needed
labels <- c("0-2cm", ">2cm-5cm", ">5cm-14cm", ">14cm")  # Category labels

# Create a new column for LifeStage based on Average.Diameter.cm
data$LifeStage <- cut(data$Average.Diameter.cm, 
                      breaks = breaks, 
                      labels = labels, 
                      right = FALSE)  # Use right=FALSE to make intervals [a, b)


```

Moorea site map
```{r}
register_google(key = "AIzaSyAJHWQg-KMSzffFNWaO1zAakoBz-klFhIg") ### use your own API

# location
Moorea<-data.frame(lon = -149.83246425684064, lat = -17.531092816791094)

#Map base
#M1<-get_map(Moorea,zoom = 12, maptype = 'satellite')

bbx <- c(left=-149.802,bottom= -17.480,right=-149.805,top=-17.475)
x <- c(bbx["left"], bbx["left"], bbx["right"], bbx["right"])
y <- c(bbx["bottom"], bbx["top"], bbx["top"], bbx["bottom"])
df <- data.frame(x, y)

sitedata <- read.csv("Data/completed.transect.site.coords.csv")
str(sitedata)

labels <- sitedata$site.name

MooreaSitemap<-ggmap(M1)+
  scalebar(x.min = -149.75, x.max = -149.79,y.min = -17.60, y.max = -17.58,
           model = 'WGS84', box.fill = c("yellow", "white"), st.color = "white",st.dist=0.3,
           location =  "bottomright", transform = TRUE, dist_unit = "km", dist = 1) +
  geom_point(data = sitedata, mapping = aes(x=long, y=lat), size=2, color="yellow")+
  geom_text(data = sitedata, aes(x=long, y=lat, label=site.name),vjust = -2,size=3, color="yellow")+
  xlab("")+
  ylab("")

MooreaSitemap


```
```{r}
#Calculate relative abundance per Site by haplotype and Life Stage


data_h <- data_h %>%
  filter(!is.na(Site) & !is.na(Haplotype))

data_h  <- data_h %>%
  filter(!Site=="NA") %>%
  filter(!Haplotype=="NA") %>%
  filter(!Sample.ID==3353)%>%
  filter(!Sample.ID==3818)%>%
  filter(!Sample.ID==3553)%>%
  filter(!Sample.ID==449)%>%
  filter(!Sample.ID==450)%>%
  filter(!Sample.ID==518)%>%
  filter(!Sample.ID==543)%>%
  filter(!Sample.ID==598)%>%
  filter(!Sample.ID==613)%>%
  filter(!Sample.ID==614)%>%
  filter(!Sample.ID==615)%>%
  filter(!Sample.ID==617)%>%
  filter(!Sample.ID==619)%>%
  filter(!Sample.ID==675)%>%
  filter(!Sample.ID==886)%>%
  filter(!Sample.ID==888)%>%
  filter(!Sample.ID==921)%>%
  filter(!Sample.ID==976)%>%
  filter(!Sample.ID==1124)%>%
  filter(!Sample.ID==1419)%>%
  filter(!Sample.ID==1838)%>%
  filter(!Sample.ID==2030)%>%
  filter(!Sample.ID==2071)%>%
  filter(!Sample.ID==2232)%>%
  filter(!Sample.ID==2459)%>%
  filter(!Sample.ID==2573)%>%
  filter(!Sample.ID==2979)%>%
  filter(!Sample.ID==3111)%>%
  filter(!Sample.ID==3122)%>%
  filter(!Sample.ID==3129)%>%
  filter(!Sample.ID==3139)%>%
  filter(!Sample.ID==3154)%>%
  filter(!Sample.ID==3258)%>%
  filter(!Sample.ID==3325)
  
  
relative_abundance.LS_h <- data_h %>%
  group_by(Site, LifeStage, Haplotype) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(Site, LifeStage) %>%
  mutate(Relative_Abundance = Count / sum(Count)) %>%
  ungroup()

relative_abundance.LS_h$ Site <- factor(relative_abundance.LS_h$Site, levels = c("S18","S15","S14", "S12", "S11","S9",
                                                                      "S6", "S5","S4", "S3", "S2","S1"))

relative_abundance.LS_h <-relative_abundance.LS_h %>%
  filter(Site %in% c("S11", "S14", "S15", "S12", "S18", "S2", "S4", "S3", "S5" , "S6", "S9", "S1"))

str(relative_abundance.LS_h)

# Calculate sample size for each Site and LifeStage
sample_sizes_h <- relative_abundance.LS_h %>%
  group_by(Site, LifeStage) %>%
  summarise(Sample_Size = sum(Count), .groups = "drop")

RA.LS.plot_h <- ggplot(relative_abundance.LS_h, aes(x = Site, y = Relative_Abundance, fill = Haplotype)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ LifeStage, ncol=5) +
  labs(x = "Site",
       y = "Relative Abundance") +
  theme_minimal() +  # Clean theme
  scale_y_continuous(labels = scales::percent) +  # Convert y-axis to percentage
  scale_fill_manual(values = color_mapping_h) +  # Set the fill colors
  theme_minimal() +
  coord_flip()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top") +
  # Add sample size text
  geom_text(data = sample_sizes_h, aes(x = Site, y = 1.1, label = Sample_Size),inherit.aes = FALSE, size = 2, vjust = 0, color = "black")
RA.LS.plot_h 

ggsave("Output/relabund.bysize_haplotype.jpg", RA.LS.plot_h , width = 6, height = 4 )

```

Calculate relative abundance per Site by Species and Life Stage
```{r}

data  <- data %>%
  filter(!Site=="NA") %>%
  filter(!Species=="NA") %>%
 # filter(!Sample.ID==3353)%>% #all good
#  filter(!Sample.ID==3818)%>% #all good
  filter(!Sample.ID==3553)%>% #no pics so no size 
  filter(!Sample.ID==449)%>% #no pics but can be fix 
  #filter(!Sample.ID==450)%>% #everything ok
  filter(!Sample.ID==518)%>% #no pics
  filter(!Sample.ID==543)%>% #no pics
  filter(!Sample.ID==598)%>% #no pics
  filter(!Sample.ID==613)%>% #no pics
  filter(!Sample.ID==614)%>% #no pics
  filter(!Sample.ID==615)%>% #no pics
  filter(!Sample.ID==617)%>% #no pics
  filter(!Sample.ID==619)%>% #no pics
  filter(!Sample.ID==675)%>% #no pics
  filter(!Sample.ID==886)%>% #no pics
  filter(!Sample.ID==888)%>% #no pics
  filter(!Sample.ID==921)%>% #no pics
  filter(!Sample.ID==976)%>% #no pics
  filter(!Sample.ID==1124)%>% #no pics
  filter(!Sample.ID==1419)%>% #no pics but chance to get it 
  filter(!Sample.ID==1838)%>% #no pics
  filter(!Sample.ID==2030)%>% #no pics
 # filter(!Sample.ID==2071)%>% #all good
  filter(!Sample.ID==2232)%>% #no pics
  filter(!Sample.ID==2459)%>% #no pics
  filter(!Sample.ID==2573)%>% #no pics
 # filter(!Sample.ID==2979)%>% #all good
  #filter(!Sample.ID==3111)%>% #all good
#  filter(!Sample.ID==3122)%>% #all good
  filter(!Sample.ID==3129)%>% #we have the pic and size but not the sequence (sequence fail)
 # filter(!Sample.ID==3139)%>% #all good
  filter(!Sample.ID==3154)%>% #no pics
#  filter(!Sample.ID==3258)%>% #all good
  filter(!Sample.ID==3325) #no pics

relative_abundance.LS <- data %>%
  group_by(Site, LifeStage, Species) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(Site, LifeStage) %>%
  mutate(Relative_Abundance = Count / sum(Count)) %>%
  ungroup()

relative_abundance.LS$Site <- factor(relative_abundance.LS$Site, levels = c("S18","S15","S14", "S12", "S11","S9",
                                                                      "S6", "S5","S4", "S3", "S2","S1"))

relative_abundance.LS <-relative_abundance.LS %>%
  filter(Site %in% c("S11", "S14", "S15", "S12", "S18", "S2", "S4", "S3", "S5" , "S6", "S9", "S1"))

str(relative_abundance.LS)

# Calculate sample size for each Site and LifeStage
sample_sizes <- relative_abundance.LS %>%
  group_by(Site, LifeStage) %>%
  summarise(Sample_Size = sum(Count), .groups = "drop")

RA.LS.plot <- ggplot(relative_abundance.LS, aes(x = Site, y = Relative_Abundance, fill = Species)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ LifeStage, ncol=5) +
  labs(x = "",
       y = "Relative Abundance") +
  theme_minimal() +  # Clean theme
  scale_y_continuous(labels = scales::percent) +  # Convert y-axis to percentage
  scale_fill_manual(values = color_mapping) +  # Set the fill colors
  theme_minimal() +
  coord_flip()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right") +
  # Add sample size text
  geom_text(data = sample_sizes, aes(x = Site, y = 1.1, label = Sample_Size),inherit.aes = FALSE, size = 1, vjust = 0, color = "black")
RA.LS.plot 

ggsave("Output/relabund.bysize.jpg", RA.LS.plot , width = 6, height = 4 )

```

Calculate relative abundance per Site by Species and Life Stage
```{r}

relative_abundance <- data %>%
  group_by(Site,Species) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(Site) %>%
  mutate(Relative_Abundance = Count / sum(Count)) %>%
  ungroup()

# Calculate sample sizes and add to relative_abundance
sample_sizes <- data %>%
  group_by(Site) %>%
  summarise(sample_size = n(), .groups = 'drop')

# Join sample sizes with relative abundance data
relative_abundance <- relative_abundance %>%
  left_join(sample_sizes, by = "Site")

relative_abundance <- relative_abundance %>%
  left_join(sample.counts, by = "Site")


relative_abundance$Site <- factor(relative_abundance$Site, levels = c("S18","S15","S14", "S12", "S11","S9",
                                                                      "S6", "S5","S4", "S3", "S2","S1"))


RA.plot <- ggplot(relative_abundance, aes(x = Site, y = Relative_Abundance, fill = Species)) +
  geom_bar(stat = "identity", position = "stack") +  # Use stacked bars
  geom_text(aes(label = sample_size, y = -0.06), 
            color = "black", 
            size = 1) +  # Adjust the size as needed
  geom_text(aes(label = total, y = -0.02), 
            color = "blue", 
            size = 1) +  # Adjust the size as needed
  labs(x = "Site",
       y = "Relative Abundance") +
  theme_minimal() +  # Clean theme
  scale_y_continuous(labels = scales::percent) +  # Convert y-axis to percentage
  scale_fill_manual(values = color_mapping) +  # Set the fill colors
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+  # Rotate x-axis labels for better visibility
  coord_flip()+
  theme(legend.position = "")
RA.plot 
```

<!-- Combine Map and Life Stage Specific Relative Abundance -->
<!-- ```{r} -->

<!-- # Combine the two plots and labels -->
<!-- p1_label <- ggdraw() + draw_plot(MooreaSitemap) + draw_label("A)", x = 0.05, y = 0.97, size = 20, fontface = "bold") -->
<!-- p2_label <- ggdraw() + draw_plot(RA.LS.plot) + draw_label("B)", x = 0.05, y = 0.97, size = 20, fontface = "bold") -->

<!-- combined_plot <- grid.arrange(p1_label, p2_label, ncol = 2) -->


<!-- # Save the combined plot -->
<!-- ggsave("Output/Poc_RA_SiteMap.pdf", plot = combined_plot, width = 14, height = 6) -->
<!-- ggsave("Output/Poc_RA_SiteMap.jpg", plot = combined_plot, width = 14, height = 6) -->


<!-- ``` -->


Combine Map and total Relative Abundance
```{r}

# Combine the two plots and labels
p1_label <- ggdraw() + draw_plot(MooreaSitemap) + draw_label("A)", x = 0.05, y = 0.97, size = 20, fontface = "bold")
p2_label <- ggdraw() + draw_plot(RA.plot) + draw_label("B)", x = 0.05, y = 0.97, size = 20, fontface = "bold")
p3_label <- ggdraw() + draw_plot(RA.LS.plot) + draw_label("C)", x = 0.05, y = 0.97, size = 20, fontface = "bold")

combined_plot.2 <- grid.arrange(p1_label, p2_label, ncol = 2)
combined_plot.3 <- grid.arrange(p1_label, p2_label, p3_label, ncol = 3)


# Save the combined plot
ggsave("Output/Poc_RA_LifeStage_SiteMap.pdf", plot = combined_plot.2, width = 20, height = 6)
ggsave("Output/Poc_RA_LifeStage_SiteMap.jpg", plot = combined_plot.2, width = 14, height = 6)


```


Density of Pocillopora
```{r}

sample.transect.counts <- all.data %>%
  group_by(Site, Transect) %>%
  summarise(total = n()) 
sample.transect.counts 

transect_counts <- all.data %>%
  group_by(Site) %>%
  summarise(num_transects = n_distinct(Transect))

transect_counts <- left_join(transect_counts, sample.counts, by="Site")

transect.area <- 50 #m2
transect_counts$Poc.density <- transect_counts$total/ (transect_counts$num_transects * transect.area)
str(transect_counts)

ggplot(transect_counts, aes(x = Poc.density, y = Site)) +
  geom_point()+
  labs(title = "Density Plot of Poc Density by Site",
       x = "Poc Density",
       y = "Density") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

```{r}
#merge the relative abundnace graph with the relative abundnace per size for each site 
#code generate with chatgpt and modifie. th plot are made with ggplot 

library(patchwork)

#combine two plots side by side 
 RA.plot + RA.LS.plot  + plot_layout(ncol = 2)

 RA.plot + RA.LS.plot + 
  plot_layout(ncol = 2) +
  plot_annotation(title = "Relative Abundance by site",
                  tag_levels = )  
combined_plot_site_size <- RA.plot + RA.LS.plot + 
  plot_layout(ncol = 2) +
  plot_annotation(title = "Relative Abundance by Species per site and size",
                  tag_levels = )

ggsave("Output/Relative_Abundance_Combined_site_size.jpg", combined_plot_site_size, width = 10, height = 5, dpi = 300)



```

#no we have a visual on what going on in the relative abundnace of the 6 sepcies of Pocillopora spp. per sizes and sites  
#our next goal is to see 
# -does site have the similar coral composition btw them?
#- Does certain species dominate the site by their size? 
#- 

```{r}
#let's make the pca to see all about our questions
# Create wide format for PCA: one row per Site, one column per Species
pca_data <- relative_abundance %>%
  select(Site, Haplotype, Relative_Abundance) %>%
  pivot_wider(names_from = Haplotype, values_from = Relative_Abundance, values_fill = 0)

# Check result
head(pca_data)

# Keep site names for labeling
site_labels <- pca_data$Site

# Prepare matrix for PCA (remove site column)
pca_matrix <- pca_data %>%
  select(-Site) %>%
  as.matrix()

# Scale the data (standard practice in PCA)
pca_scaled <- scale(pca_matrix)

# Run PCA
pca_result <- prcomp(pca_scaled, center = TRUE, scale. = TRUE)

# View summary
summary(pca_result)


```

```{r}
#visualization of the PCA plot 
# Create dataframe for plotting
pca_scores <- as.data.frame(pca_result$x)
pca_scores$Site <- site_labels

# Plot PCA
ggplot(pca_scores, aes(x = PC1, y = PC2, label = Site)) +
  geom_point(aes(color = Site), size = 4) +
  geom_text(vjust = -1.5, size = 3.5) +
  labs(title = "PCA of Coral Species Composition by Site",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "% variance)")) +
  theme_minimal()

```
```{r}
# Hierarchical clustering
dist_matrix <- dist(pca_scaled)
hc <- hclust(dist_matrix)

# Dendrogram
plot(hc, labels = site_labels, main = "Clustering of Sites based on Species Composition")

# K-means on PC1 and PC2
set.seed(123)
kmeans_result <- kmeans(pca_scores[,1:2], centers = 3)

# Add cluster to PCA plot
pca_scores$Cluster <- as.factor(kmeans_result$cluster)

ggplot(pca_scores, aes(x = PC1, y = PC2, color = Cluster, label = Site)) +
  geom_point(size = 4) +
  geom_text(vjust = -1.2) +
  labs(title = "PCA with K-Means Clustering") +
  theme_minimal()

```
```{r}
#running the permanova of the plot 
library(vegan)

# Use original unscaled matrix
adonis_result <- adonis2(pca_matrix ~ site_labels, method = "euclidean")

adonis_result

```
```{r}
#now i want to try to see the answer of this questions: Does certain species dominate the site by their size? 
ggplot(relative_abundance.LS, aes(x = LifeStage, y = Relative_Abundance, fill = Species)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ Site, ncol = 4) +
  labs(title = "Relative Abundance of Species by Size Class and Site",
       x = "Size Class",
       y = "Relative Abundance") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = color_mapping) +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))

#to look at a different way (table)
dominant_species_table <- relative_abundance.LS %>%
  group_by(Site, LifeStage) %>%
  slice_max(Relative_Abundance, n = 1, with_ties = FALSE) %>%
  select(Site, LifeStage, Species, Relative_Abundance)

dominant_species_table

```
```{r}
#filtering the dominant species that make up more than 70% of a size class at a site
dominant_strong <- dominant_species_table %>%
  filter(Relative_Abundance > 0.7)

dominant_strong

#this show us only the species that are truly dominating 
```

#site frequency plot 
```{r}
#ggplot(data_h, aes(x = Average.Diameter.cm, y = Count, color = Species, group = Haplotype)) +
#  geom_line(size = 1.2) +
#  facet_wrap(~ Site, scales = "free_y") +
#  theme_minimal() +
#  labs(title = "Size Class Trends per Species Across Sites",
#       y = "Count", x = "Size Class") +
#  scale_color_manual(values = color_mapping) +
#  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create named color vector for mapping
color_mapping_h <- setNames(cols.haplotype$Color, cols.haplotype$Haplotype)


# Plot with custom colors
frequency <- ggplot(data_h, aes(x = Average.Diameter.cm, colour = Haplotype, fill = Haplotype)) +
  geom_freqpoly(bins = 10) +
  facet_wrap(~Site, ncol = 3) +
  scale_color_manual(values = color_mapping_h) +  # apply your color mapping
  scale_fill_manual(values = color_mapping_h) +   # apply for fill (if needed)
  theme_bw()

frequency
ggsave("Output/Relative_Abundance_haplotype_frequency.jpg", frequency, width = 10, height = 5, dpi = 300)

#pretty version of the graph bellow 
frequency_p <- ggplot(data_h, aes(x = Average.Diameter.cm, colour = Haplotype)) +
  geom_freqpoly(bins = 10, size = 1.2) +
  xlim(0, NA)+ # Thicker lines for visibility
  facet_wrap(~Site, ncol = 3, scales = "free_y") +  # Adjust y-axis per site
  scale_color_manual(values = color_mapping_h, name = "Haplotype") +
  labs(
    title = "Size Frequency of Pocillopora Haplotypes by Site",
    subtitle = "Based on Average Colony Diameter",
    x = "Average Diameter (cm)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 10),  # Facet titles
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 9),
    panel.grid.minor = element_blank()  # Clean up minor grid lines
  ) 

frequency_p
ggsave("Output/Relative_Abundance_haplotype_frequency_p.jpg", frequency_p, width = 10, height = 5, dpi = 300)

```


```{r}
#Prepare species-by-site matrix
relative_abundance <- data %>%
  group_by(Site, Species) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  group_by(Site) %>%
  mutate(Relative_Abundance = Count / sum(Count)) %>%
  ungroup()

library(tidyr)

species_matrix <- relative_abundance %>%
  select(Site, Species, Relative_Abundance) %>%
  pivot_wider(names_from = Species, values_from = Relative_Abundance, values_fill = 0) %>%
  column_to_rownames("Site")  # set rownames to Site for PCA

pca_result <- prcomp(species_matrix, scale. = TRUE)  # scale = TRUE standardizes variables
summary(pca_result)

library(ggfortify)

autoplot(pca_result,
         data = data.frame(Site = rownames(species_matrix)),
         label = TRUE, label.size = 4,
         colour = 'Site') +
  theme_minimal() +
  labs(title = "PCA of Pocillopora Species Composition by Site")

library(factoextra)

fviz_pca_biplot(pca_result,
                repel = TRUE,
                label = "var", # or "none", "var", "ind"
                col.var = "black",
                col.ind = "steelblue",
                geom.ind = "point") +
  labs(title = "PCA Biplot: Species Composition by Site")

#adding at the analysis site + Sizeclass 
size_species_matrix <- relative_abundance.LS %>%
  group_by(Site, LifeStage, Species) %>%
  summarise(Relative_Abundance = sum(Relative_Abundance), .groups = "drop") %>%
  unite("Group", Site, LifeStage, sep = "_") %>%
  pivot_wider(names_from = Species, values_from = Relative_Abundance, values_fill = 0) %>%
  column_to_rownames("Group")

pca_result_size <- prcomp(size_species_matrix, scale. = TRUE)
fviz_pca_ind(pca_result_size, repel = TRUE)

```
#site frequency 
```{r}

```


