---
title: "Size recruit classification"
author: "Pierrick"
date: '2024-01-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Install and load packages

library("ggplot2")
library("ggpubr")
library("tidyverse")
library("viridisLite")
library("dplyr")

```

load data for size recruit scale
```{r}
data_scale<- read.csv("/Users/pierrickharnay/Dropbox/MyProjects/Pocillopora-Lagoon-Abundance/Data/POC_ID_RA_list_size.csv", sep=";", header = TRUE, na.strings = "NA")
colnames(data_scale)
class(data_scale)
str(data_scale)

range(data_scale$cm_measure)
range(na.omit(data_scale$cm_measure))
```

########### Calculate a confidence interval for cm ###################
```{r}
##### Step 1: Calculate the mean #####
data("data_scale")
sample.mean <- mean(data_scale$cm_measure)
print(sample.mean)
#1] 0.99998

##### Step 2: Calculate the standard error of the mean #####  
sample.n <- length(data_scale$cm_measure)
sample.sd <- sd(data_scale$cm_measure)
sample.se <- sample.sd/sqrt(sample.n)
print(sample.se)
#[1] 0.001238

#####  Step 3: Find the t-score that corresponds to the confidence level #####
alpha = 0.05
degrees.freedom = sample.n - 1
t.score = qt(p=alpha/2, df=degrees.freedom,lower.tail=F)
print(t.score)

#####  Step 4. Calculate the margin of error and construct the confidence interval #####  
margin.error <- t.score * sample.se
  lower.bound <- sample.mean - margin.error
  upper.bound <- sample.mean + margin.error
  print(c(lower.bound,upper.bound))
#[1] 0.9974921 1.0024679 
  
##### linear regression ##### 
##### Mean and standard error #####  
l.model <- lm(cm_measure ~ 1, data_scale)

#####  Calculate the confidence interval #####  
confint(l.model, level=0.95)

#  2.5 %   97.5 %
#(Intercept) 0.9974921 1.002468
  
########### Calculate a confidence interval for mm ###################
##### Step 1: Calculate the mean #####
data("data_scale")
sample.mean <- mean(data_scale$mm_measure)
print(sample.mean)
#[1] 0.10906

##### Step 2: Calculate the standard error of the mean #####  
sample.n <- length(data_scale$mm_measure)
sample.sd <- sd(data_scale$mm_measure)
sample.se <- sample.sd/sqrt(sample.n)
print(sample.se)
#1] 0.001166684

#####  Step 3: Find the t-score that corresponds to the confidence level #####
alpha = 0.05
degrees.freedom = sample.n - 1
t.score = qt(p=alpha/2, df=degrees.freedom,lower.tail=F)
print(t.score)

#####  Step 4. Calculate the margin of error and construct the confidence interval #####  
margin.error <- t.score * sample.se
  lower.bound <- sample.mean - margin.error
  upper.bound <- sample.mean + margin.error
  print(c(lower.bound,upper.bound))
#[1] 0.1067155 0.1114045

##### linear regression ##### 
##### Mean and standard error #####  
l.model <- lm(mm_measure ~ 1, data_scale)

#####  Calculate the confidence interval #####  
confint(l.model, level=0.95)
#2.5 %    97.5 %
#(Intercept) 0.1067155 0.1114045  


```
load the data 
```{r}
data_size<- read.csv("Data/POC_ID_RA_list_size.csv", sep=",", header = TRUE, na.strings = "NA")
colnames(data_size)
class(data_size)
str(data_size)

range(data_size$Recruit_size_cm2)
range(na.omit(data_size$Recruit_size_cm2))

#calculate the diameter assuming the shape is a circle
data_size$Recruit_diameter_cm <- (2*(sqrt(pi/data_size$Recruit_size_cm2)))/100
```

Size Class definitions
```{r}


#Class 1 0 - 10mm in diameter = 
#Class 2 >10 - 40mm in diameter = 
#Class 3 >40 - 60mm in diameter = 
#Class 4 >60 - 100mm in diameter =

#set the size class diameter cm2
c1 <- 1
c2 <- 4
c3 <- 6
c4 <- 10

#Class 1 0 - 1cm in diameter Area 0 - 0.7854 cm2
c1A <- pi*(c1/2)^2

#Class 2 1 - 4cm in diameter Area >0.7854 - 12.56637 cm2
c2A <- pi*(c2/2)^2

#Class 3 4 - 6cm in diameter Area >12.56637 - 28.27433 cm2
c3A <- pi*(c3/2)^2

#Class 4 6 - 10cm in diameter Area >28.27433 - 78.53982 cm2
c4A <- pi*(c4/2)^2

#Class 4 6 - 10cm in diameter Area >78.54 cm2
c5A <- pi*(c4/2)^2

```

#Load Data and QC
```{r}
#data_size_cm2 <- read.csv("/Users/pierrickharnay/Dropbox/MyProjects/Pocillopora-Lagoon-Abundance/Data/POC_RA_Size.csv", sep=";", header = TRUE, na.strings = "NA")

data_size_cm2 <- data_size[c(1:2056),]

#Sanity checks of data
unique(data_size_cm2$Recruit_size_cm2)
unique(data_size_cm2$Site)
unique(data_size_cm2$Transect)
unique(data_size_cm2$Number)

#change NA to indicate NA is adult size or >10cm in diameter
data_size_cm2 <- data_size_cm2 %>%
  mutate(Recruit_size_cm2 = if_else(is.na(Recruit_size_cm2), 78.54, Recruit_size_cm2))

#remove 0. 0 indicates the photograph was not able to be analyzed
data_size_cm2 <- data_size_cm2 %>%
  filter(Recruit_size_cm2>0)
```

Assign size classes
```{r}

#assign size class based on area
data_size_cm2 <-data_size_cm2 %>% 
  mutate(Size.Class = case_when(Recruit_size_cm2 < c1A ~ "Class1",
         Recruit_size_cm2 > c1A & Recruit_size_cm2 < c2A ~ "Class2",
         Recruit_size_cm2 > c2A & Recruit_size_cm2 < c3A ~ "Class3",
         Recruit_size_cm2 > c3A & Recruit_size_cm2 < c4A ~ "Class4",
         Recruit_size_cm2 > c4A ~ "Class5",
         TRUE ~ "ERROR"))

# Calculate the number and frequency of samples for each class on each transect and site
x <- data_size_cm2 %>% 
  group_by(Site, Transect, Size.Class) %>% 
  summarise(n=n()) 

#plot number of colonies by size class
class.freq <- x %>% ggplot(aes(Size.Class, n, fill = Size.Class)) +
  geom_col()+
  scale_fill_manual(values = c("white","lightgray","gray","darkgray", "black"))+
  facet_wrap(vars(Site,Transect), ncol = 6)+
  theme_bw()+
  theme(axis.text.x=element_text(angle = 90, hjust = 0))+
  ylab("Number of Colonies in each size class")+
  xlab("Size Class")
class.freq 

ggsave("Output/number_per_class.pdf", class.freq, height=10, width =6 )

```

#make a colum by group of size zith categorie without adults 
```{r}


# Définir les intervalles de taille pour la découpe
intervals <- c(0, 1, 4, 6, 10)

# Créer une nouvelle colonne 'groupe_taille' en fonction des intervalles
data_size_cm2$Groups <- cut(data_size_cm2$Recruit_size_cm2, breaks = intervals, labels = c("Recruit", "Juvenille", "teenager", "Young adult"), include.lowest = TRUE)

``` 


#Calculate relative abundance of group of size
```{r}

Rel.Abund <- data_size_cm2 %>%
  group_by(Site, Transect, Groups, Recruit_size_cm2) %>%
  tally() %>% 
  mutate(freq = prop.table(n)) %>% 
  mutate(tot = sum(n))

sample.sizes <- Rel.Abund %>%
 filter(row_number() %% 2 == 1)

Rel.Abund$Site <- factor(Rel.Abund$Site, levels=c("S1", "S2", "S3", "S4", "S5", "S11", "S15", "S9", "S18"))
Rel.Abund$Site

#"S4"  "S3"  "S5"  "S1"  "S15" "S2"  "S11"

data_size_cm2 <- data_size_cm2 %>%
  filter(Recruit_size_cm2!="NA")
data_size_cm2 <- data_size_cm2 %>%
  filter(Recruit_size_cm2!="0")
```

# identify sample sizes for each site
```{r}

S1.sample.sizes <- sample.sizes %>%
  filter(Site=="S1")

S2.sample.sizes <- sample.sizes %>%
  filter(Site=="S2")

S3.sample.sizes <- sample.sizes %>%
  filter(Site=="S3")

S4.sample.sizes <- sample.sizes %>%
  filter(Site=="S4")

S5.sample.sizes <- sample.sizes %>%
  filter(Site=="S5")

S11.sample.sizes <- sample.sizes %>%
  filter(Site=="S11")

S15.sample.sizes <- sample.sizes %>%
  filter(Site=="S15")

S9.sample.sizes <- sample.sizes %>%
  filter(Site=="S9")

S18.sample.sizes <- sample.sizes %>%
  filter(Site=="S18")

```


#Plot by Transect by Site
```{r}

S1.plot_size <- Rel.Abund %>%
  filter(Site=="S1") %>%
ggplot(., aes(x = Transect, y = freq, fill=Groups)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey", "purple", "green", "pink"))+
    geom_text( data = S1.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Groups") +
  theme_classic() + 
  theme_(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1, size=10))+
  theme(axis.text.y = element_text(size=5))+
  theme(axis.title.y = element_text(size=5))+
  theme(axis.line=element_line()) 

S2.plot_size <- Rel.Abund %>%
  filter(Site=="S2") %>%
ggplot(., aes(x = Transect, y = freq, fill=Groups)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey", "purple", "green", "pink"))+
    geom_text( data = S2.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Groups") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=20))+
  theme(axis.text.y = element_text(size=20))+
  theme(axis.title.y = element_text(size=28))+
  theme(axis.line=element_line()) 

S3.plot <- Rel.Abund %>%
  filter(Site=="S3") %>%
ggplot(., aes(x = Transect, y = freq, fill=Stage)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey"))+
    geom_text( data = S3.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Stage") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=20))+
  theme(axis.text.y = element_text(size=20))+
  theme(axis.title.y = element_text(size=28))+
  theme(axis.line=element_line()) 

S4.plot <- Rel.Abund %>%
  filter(Site=="S4") %>%
ggplot(., aes(x = Transect, y = freq, fill=Stage)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey"))+
    geom_text( data = S4.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Stage") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=20))+
  theme(axis.text.y = element_text(size=20))+
  theme(axis.title.y = element_text(size=28))+
  theme(axis.line=element_line()) 

S5.plot <- Rel.Abund %>%
  filter(Site=="S5") %>%
ggplot(., aes(x = Transect, y = freq, fill=Stage)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey"))+
    geom_text( data = S5.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Stage") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=20))+
  theme(axis.text.y = element_text(size=20))+
  theme(axis.title.y = element_text(size=28))+
  theme(axis.line=element_line()) 

S11.plot.data <- Rel.Abund %>%
  filter(Site=="S11") 
  
extra.11 <-  data.frame (Site  = c("S11","S11","S11","S11","S11","S11","S11","S11"),
                  Transect = c("T3", "T3","T4", "T4","T5", "T5","T6", "T6"),
                  group = c("S11 T3","S11 T3","S11 T4","S11 T4","S11 T5","S11 T5","S11 T6","S11 T6"), 
                  Stage = c("Adult", "Recruit","Adult", "Recruit","Adult", "Recruit","Adult", "Recruit"),
                  n= c(0,0,0,0,0,0,0,0),
                  freq = c(0,0,0,0,0,0,0,0),
                  tot = c(0,0,0,0,0,0,0,0))

S11.plot.data <- rbind(S11.plot.data, extra.11)

S11.plot <- S11.plot.data %>%
ggplot(., aes(x = Transect, y = freq, fill=Stage)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey"))+
  scale_x_discrete(drop = FALSE) +
    geom_text( data = S11.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Stage") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=20))+
  theme(axis.text.y = element_text(size=20))+
  theme(axis.title.y = element_text(size=28))+
  theme(axis.line=element_line()) 

S15.plot.data <- Rel.Abund %>%
  filter(Site=="S15") 
  
extra.15 <-  data.frame (Site  = c("S15","S15","S15","S15"),
                  Transect = c("T5", "T5","T6", "T6"),
                  group = c("S15 T5","S15 T5","S15 T6","S15 T6"), 
                  Stage = c("Adult", "Recruit","Adult", "Recruit"),
                  n= c(0,0,0,0),
                  freq = c(0,0,0,0),
                  tot = c(0,0,0,0))

S15.plot.data <- rbind(S15.plot.data, extra.11)
  
  
S15.plot <- S15.plot.data %>%
ggplot(., aes(x = Transect, y = freq, fill=Stage)) +
  geom_col() +
  scale_fill_manual(values = c("black", "grey"))+
  scale_x_discrete(drop = FALSE) +
    geom_text( data = S15.sample.sizes,aes(x = Transect, y = 1.05, label = tot)) +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Stage") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=20))+
  theme(axis.text.y = element_text(size=20))+
  theme(axis.title.y = element_text(size=28))+
  theme(axis.line=element_line()) 

ggsave("/Users/pierrickharnay/Dropbox/MyProjects/Pocillopora-Lagoon-Abundance/Output/S1_RA_size.jpg", S1.plot_size, height=6, width=9)
ggsave("/Users/pierrickharnay/Dropbox/MyProjects/Pocillopora-Lagoon-Abundance/Output/S2_RA_size.jpg", S2.plot_size, height=4, width=5)
ggsave("Output/S3_RA.jpg", S3.plot, height=4, width=5)
ggsave("Output/S4_RA.jpg", S4.plot, height=4, width=5)
ggsave("Output/S5_RA.jpg", S5.plot, height=4, width=5)
ggsave("Output/S11_RA.jpg", S11.plot, height=4, width=5)
ggsave("Output/S15_RA.jpg", S15.plot, height=4, width=5)

```

#calculate the density of all Pocillopora per site
```{r}

#50m x 0.5m is the area of a transect
Rel.Abund.Site <- data %>%
  group_by(Site, Stage) %>%
  tally() %>% 
  mutate(freq = prop.table(n)) %>% 
  mutate(tot = sum(n))

Rel.Abund.Site$Site <- factor(Rel.Abund.Site$Site, levels=c("S1", "S2", "S3", "S4", "S5", "S11", "S15"))
Rel.Abund.Site$Site

site.sample.sizes <- Rel.Abund.Site %>%
 filter(row_number() %% 2 == 1)

#calculate the number of transects to get density per site
density <- data %>%
  group_by(Site, Transect) %>%
  tally() 

site.density <- density %>%
  group_by(Site) %>%
  summarise(total=sum(n), 
            transect.number = length(unique(Transect))) %>%
  mutate(density = total/(transect.number*25))

scale_factor <- max(site.density$density) /max(Rel.Abund.Site$freq) 


Site.plot <- Rel.Abund.Site %>%
ggplot(., aes(x = Site, y = freq)) +
  geom_col(aes(x = Site, y = freq, fill=Stage)) +
  scale_fill_manual(values = c("black", "grey"))+
  geom_point(data =site.density, aes(x=Site, y=density/scale_factor), color = "red", size=3) +
  #scale_y_continuous("Relative Abundance", sec.axis = dup_axis(name = "Density (corals/m2)")) +
  scale_y_continuous("Relative Abundance", sec.axis=sec_axis(~.*scale_factor, name="Density (corals/m2)")) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=18),
        axis.text.y = element_text(size=18),
        axis.title.y = element_text(size=24),
        axis.title.x = element_blank(),
        axis.line=element_line()) 

ggsave("Output/Site_RA.jpg", Site.plot, height=4, width=5)

```
